---
title: "Identification of Pathogenic Immune Cell Subsets Associated with Checkpoint Inhibitor-induced Myocarditis"
output: pdf_document
---

 - data reproduction of https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9397491/
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries

```{r}

library(Seurat)
library(SeuratDisk)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)

```

# Set theme

```{r}

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_rect(colour="black", fill = NA, size = 1),
          axis.text = element_text(colour="black"),
          axis.ticks = element_line(colour="black"),
          axis.text.x = element_text(angle=0),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 7)))

```

# Read in scRNAseq RDS

```{r}

immune.combined <- readRDS("../saved_data/final_seurat.rds")

```

# Read in scTCRseq, process to be added as metadata to scRNAseq

```{r}

dehashed_combined_tcr_seq <- read.csv("../saved_data/dehashed_tcr_seq.csv")
  
suffix_lookup <- tibble(
  Patient_ID = c("B1", "A10", "MCE12", "B10", "MCE13", "B12", "MCE7", "MCE6", "A8", "B11", "A9", "A6", "B9", "B13", "A5", "B14", "B2", "A1", "B3", "A2", "A3", "MCE1", "MCE2", "MCL1", "MCL3"),
  suffix = c("10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25", "1", "2", "3", "4", "5", "6", "7", "8", "9")
)

# Merge the lookup table to add the suffix column to the main data frame
dehashed_combined_tcr_seq <- dehashed_combined_tcr_seq %>%
  left_join(suffix_lookup, by = "Patient_ID") %>%
  mutate(barcode = paste(barcode, suffix, sep = "_")) %>%
  filter(chain != "Multi") %>%
  # pivot to combine 1 alpha and 1 beta (or 0 and 1) barcodes together
  select(-X) %>%
    # if a cell had two or more qualified chains of the same type, only that chain with the highest UMI count was qualified and retained
  group_by(barcode, chain) %>%
  arrange(-umis) %>%
  slice(1) %>%
  ungroup() %>%
  pivot_wider(id_cols = c(barcode, Patient_ID, irAE, suffix), names_from = chain, values_from = -c(barcode, Patient_ID, irAE, suffix))

dehashed_combined_tcr_seq <- as.data.frame(dehashed_combined_tcr_seq)

# Set row names based on the "Barcode" column
rownames(dehashed_combined_tcr_seq) <- dehashed_combined_tcr_seq$barcode

dehashed_combined_tcr_seq <- dehashed_combined_tcr_seq %>% select(-barcode)

dehashed_combined_tcr_seq

```

# Add TCRseq metadata to RNAseq

```{r}

rna_and_tcr <- AddMetaData(object = immune.combined, metadata = dehashed_combined_tcr_seq)

```

# Look at just T cells

```{r}

T_cells <- rna_and_tcr[, !(is.na(rna_and_tcr@meta.data$cdr3_TRA) | is.na(rna_and_tcr@meta.data$cdr3_TRB)) & rna_and_tcr@meta.data$predicted.celltype.l1 %in% c("CD4 T", "CD8 T", "other T") & rna_and_tcr@meta.data$predicted.celltype.l2 %in% c("CD8 TEM", "CD4 TEM", "CD8 TCM", "CD4 TCM", "Treg", "CD4 Naive", "CD8 Naive",  "dnT", "CD4 Proliferating", "CD8 Proliferating", "MAIT", "CD4 CTL", "CD8 CTL")]

```

# Re-do dimensional reduction, clustering

```{r}

T_cells <- RunPCA(T_cells, npcs = 30, verbose = FALSE)
T_cells <- RunUMAP(T_cells, reduction = "pca", dims = 1:30)
T_cells <- FindNeighbors(T_cells, reduction = "pca", dims = 1:30)
T_cells <- FindClusters(T_cells, resolution = 3)

```

# Read in reference to use for mapping

```{r}

reference <- LoadH5Seurat("../raw_data/pbmc_multimodal.h5seurat")

```

# Find reference transfer anchors

```{r}

anchors <- FindTransferAnchors(
  reference = reference,
  query = T_cells,
  normalization.method = "SCT",
  reference.reduction = "pca",
  dims = 1:50
)

```

# Map reference onto my object

```{r}

T_cells <- MapQuery(
  anchorset = anchors,
  query = T_cells,
  reference = reference,
  refdata = list(
    celltype.l1 = "celltype.l1",
    celltype.l2 = "celltype.l2",
    predicted_ADT = "ADT"
  ),
  reference.reduction = "pca", 
  reduction.model = "wnn.umap"
)

```

# Plot mapping results

```{r}

p1 = DimPlot(T_cells, reduction = "umap", group.by = "predicted.celltype.l1", label = TRUE, label.size = 3, repel = TRUE) + NoLegend()

p2 = DimPlot(T_cells, reduction = "umap", group.by = "predicted.celltype.l2", label = TRUE, label.size = 3 , repel = TRUE) + NoLegend()

p1 + p2

```

# Relook just at T cells again

```{r}

just_T_cells <- T_cells[, !(is.na(T_cells@meta.data$cdr3_TRA) | is.na(T_cells@meta.data$cdr3_TRB)) & T_cells@meta.data$predicted.celltype.l1 %in% c("CD4 T", "CD8 T", "other T") & T_cells@meta.data$predicted.celltype.l2 %in% c("CD8 TEM", "CD4 TEM", "CD8 TCM", "CD4 TCM", "Treg", "CD4 Naive", "CD8 Naive",  "dnT", "CD4 Proliferating", "CD8 Proliferating", "MAIT", "CD4 CTL", "CD8 CTL")]

just_T_cells <- RunPCA(just_T_cells, npcs = 30, verbose = FALSE)
just_T_cells <- RunUMAP(just_T_cells, reduction = "pca", dims = 1:30)
just_T_cells <- FindNeighbors(just_T_cells, reduction = "pca", dims = 1:30)
just_T_cells <- FindClusters(just_T_cells, resolution = 3)
anchors <- FindTransferAnchors(
  reference = reference,
  query = just_T_cells,
  normalization.method = "SCT",
  reference.reduction = "pca",
  dims = 1:50
)
T_cells <- MapQuery(
  anchorset = anchors,
  query = just_T_cells,
  reference = reference,
  refdata = list(
    celltype.l1 = "celltype.l1",
    celltype.l2 = "celltype.l2",
    predicted_ADT = "ADT"
  ),
  reference.reduction = "pca", 
  reduction.model = "wnn.umap"
)

```

# Plot cell type assignments

```{r}

p1 = DimPlot(just_T_cells, reduction = "umap", group.by = "predicted.celltype.l1", label = TRUE, label.size = 3, repel = TRUE) + NoLegend()

p2 = DimPlot(just_T_cells, reduction = "umap", group.by = "predicted.celltype.l2", label = TRUE, label.size = 3 , repel = TRUE) + NoLegend()

p1 + p2

```

# Save RDS

```{r}

saveRDS(just_T_cells, "../saved_data/final_seurat_T_cells.rds")

```

# Map clonal relations

```{r}

public_df <- just_T_cells@meta.data %>%
  group_by(cdr3_TRA, cdr3_TRB) %>%
  # just use CDR3 for now...? can also try with V(D)JC and see how different it is...
  # need to fix early/late patient still
  summarize(count = n()) %>%
  filter(count > 1) %>%
  mutate(joined_cdr3 = paste(cdr3_TRA, cdr3_TRB, sep = ""))
  
public_df

clonal_relations <- just_T_cells@meta.data %>%
  mutate(joined_cdr3 = paste(cdr3_TRA, cdr3_TRB, sep = "")) %>%
  mutate(public = if_else(joined_cdr3 %in% public_df$joined_cdr3, "public", "private"))

clonal_relations

# Add metadata of publicness back

just_T_cells <- AddMetaData(object = just_T_cells, metadata = clonal_relations)

```

# Airline plot


```{r}

# Subset public cells
public_cells <- subset(just_T_cells, subset = public == "public")

# Extract UMAP coordinates for public cells
umap_coordinates <- Embeddings(public_cells, "umap")

# Plot UMAP
p2 <- DimPlot(just_T_cells, reduction = "umap", group.by = "predicted.celltype.l2", label = TRUE, label.size = 3, repel = TRUE) + NoLegend()

# Loop through unique flavors
for (flavor in unique(clonal_relations$joined_cdr3)) {
  # Subset public cells of the current flavor
  flavor_cells <- WhichCells(public_cells, cells = row.names(public_cells)[public_cells$joined_cdr3 == flavor])

  # Extract UMAP coordinates for cells of the current flavor
  flavor_umap_coords <- umap_coordinates[row.names(flavor_cells), ]
  
  # Add lines connecting cells of the same flavor
  if (nrow(flavor_umap_coords) > 1) {
    lines_data <- data.frame(
      x = flavor_umap_coords[1:(nrow(flavor_umap_coords) - 1), 1],
      y = flavor_umap_coords[1:(nrow(flavor_umap_coords) - 1), 2],
      xend = flavor_umap_coords[2:nrow(flavor_umap_coords), 1],
      yend = flavor_umap_coords[2:nrow(flavor_umap_coords), 2]
    )
    
    p2 <- p2 + geom_segment(data = lines_data, aes(x = x, y = y, xend = xend, yend = yend), color = "black", size = 0.5)
  }
}

# Display the plot
print(p2)

```