---
title: "Identification of Pathogenic Immune Cell Subsets Associated with Checkpoint Inhibitor-induced Myocarditis"
output: pdf_document
---

 - data reproduction of https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9397491/
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries

```{r}

library(Seurat)
library(Matrix)
library(ggplot2)
library(dplyr)

```

# Set theme

```{r}

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_rect(colour="black", fill = NA, size = 1),
          axis.text = element_text(colour="black"),
          axis.ticks = element_line(colour="black"),
          axis.text.x = element_text(angle=0),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 7)))

```

# Read in RDS of QC filtered data

```{r}

heart_t_cells_filtered <- readRDS("../saved_data/qc_filtered.rds")

```

# Split data by sample before integrating by finding anchors

```{r}

# split the dataset into a list of seurat objects (for each patient sample)
heart_t_cells_filtered.list <- SplitObject(heart_t_cells_filtered, split.by = "orig.ident")

# normalize and identify variable features for each dataset independently
heart_t_cells_filtered.list <- lapply(X = heart_t_cells_filtered.list, FUN = SCTransform, method = "glmGamPoi")

# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = heart_t_cells_filtered.list, nfeatures = 3000)

# prep step
heart_t_cells_filtered.list <- PrepSCTIntegration(object.list = heart_t_cells_filtered.list, anchor.features = features)

# run PCA to allow rpca use later
heart_t_cells_filtered.list <- lapply(X = heart_t_cells_filtered.list, FUN = RunPCA, features = features)

```

# Randomly select a single sample per group to use as a reference

```{r}

# all samples have been downsampled to the same number of cells, so I can choose randomly (with seed)

# Set a random seed for reproducibility
set.seed(123)

# Extract the starting letters from the names
starting_letters <- substr(names(heart_t_cells_filtered.list), 1, 1)

# Create a data frame with Seurat object names and their starting letters
seurat_info <- data.frame(name = names(heart_t_cells_filtered.list), starting_letter = starting_letters)

# Randomly sample one Seurat object from each starting letter group
randomly_selected_seurat <- seurat_info %>%
  group_by(starting_letter) %>%
  sample_n(1) %>%
  ungroup() %>%
  select(name)

randomly_selected_seurat

```

# Find integration anchors using the above single sample/group as references

```{r}

# first integrates these references, then aligns other samples within each group to that integrated reference
# this command creates an 'integrated' data assay

immune.anchors <- FindIntegrationAnchors(object.list = heart_t_cells_filtered.list, reference = match(heart_t_cells_filtered.list, randomly_selected_seurat$name), normalization.method = "SCT", anchor.features = features, reduction = "rpca", k.anchor = 10, dims = 1:30)

```

# Integrate the data

```{r}

immune.combined <- IntegrateData(anchorset = immune.anchors, normalization.method = "SCT")

```

# Specify integration, scale data

```{r}

# specify that we will perform downstream analysis on the corrected data note that the
# original unmodified data still resides in the 'RNA' assay
DefaultAssay(immune.combined) <- "integrated"

# scale data
immune.combined <- ScaleData(immune.combined, verbose = FALSE)

```

# PCA using 2000 most variable genes and top 30 PCs

```{r}

immune.combined <- RunPCA(immune.combined, npcs = 30, verbose = FALSE)

```

# UMAP using 2000 most variable genes and top 30 PCs

```{r}

immune.combined <- RunUMAP(immune.combined, reduction = "pca", dims = 1:30)

```

# Find clusters

```{r}

immune.combined <- FindNeighbors(immune.combined, reduction = "pca", dims = 1:30)
immune.combined <- FindClusters(immune.combined, resolution = 3)

```

# Visualize UMAP by patient ID

```{r}

DimPlot(immune.combined, reduction = "umap", group.by = "orig.ident")

ggsave("../figures/umap_patient_id.svg")

```

# Visualize UMAP by cluster

```{r}

DimPlot(immune.combined, reduction = "umap", label = TRUE, repel = TRUE)

ggsave("../figures/umap_cluster.svg")

```

# Save RDS

```{r}

saveRDS(immune.combined, "../saved_data/transformed_reduced.rds")

```