---
title: "Identification of Pathogenic Immune Cell Subsets Associated with Checkpoint Inhibitor-induced Myocarditis"
output: pdf_document
---

 - data mining of https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9397491/
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries

```{r}

library(ggplot2)
library(dplyr)
library(tidyr)
library(VennDiagram)
library(scales)

```

# Set theme

```{r}

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_rect(colour="black", fill = NA, size = 1),
          axis.text = element_text(colour="black"),
          axis.ticks = element_line(colour="black"),
          axis.text.x = element_text(angle=0),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 7)))

```

# Read in data

```{r}

feature_df <- read.csv("../saved_data/tcr_seq_features.csv")

feature_df

```

# Import pgen data, join with other data

```{r}

igor_alpha <- read.csv("../saved_data/igor_output_alpha.csv") %>%
  dplyr::rename(TRA_pgen = pgen) %>%
  filter(!(is.na(TRA_pgen)),
         TRA_pgen != "NaN")

combined_data <- read.csv("../saved_data/igor_output_beta.csv") %>%
  dplyr::rename(TRB_pgen = pgen) %>%
  filter(!(is.na(TRB_pgen)),
         TRB_pgen != "NaN") %>%
  full_join(igor_alpha, by = c("TCR_name", "TRA_nt", "TRB_nt", "TRA_aa", "TRB_aa", "TRAV", "TRAJ", "TRA_CDR3", "TRBV", "TRBJ", "TRB_CDR3", "TRAC", "TRBC", "TRA_leader", "TRB_leader", "Linker", "Link_order", "TRA_5_prime_seq", "TRA_3_prime_seq", "TRB_5_prime_seq", "TRB_3_prime_seq", "Linked_nt", "Linked_aa", "Warnings.Errors")) %>%
  select(TCR_name, TRA_CDR3, TRB_CDR3, TRA_pgen, TRB_pgen) %>%
  mutate(TRA_pgen = as.character(TRA_pgen),
         TRB_pgen = as.character(TRB_pgen)) %>%
  pivot_longer(cols = matches("^(TRA|TRB)"),
               names_to = c("Chain", ".value"),
               names_pattern = "^(TRA|TRB)_(.*)$") %>%
  dplyr::rename(barcode = TCR_name,
         cdr3 = CDR3) %>%
  mutate(pgen = as.numeric(pgen)) %>%
  right_join(feature_df, by = c("barcode", "Chain", "cdr3"))

combined_data

```
  
# Examine patient sampling depth bias

```{r}

seq_depth <- combined_data %>%
  group_by(Patient_ID, irAE) %>%
  summarize(seq_depth = n()) %>%
  arrange(seq_depth) %>%
  print()

seq_depth

# Choose 324 seqs from non B10 samples to downsample all to 324

```

# Downsample

```{r}

desired_rows_per_patient <- 324/2
# /2 to get equal numbers of each chain (TRA and TRB)

set.seed(123)

# Downsample the dataframe
downsampled_df <- combined_data %>%
  group_by(Patient_ID, Chain) %>%
  sample_n(desired_rows_per_patient, replace = FALSE) %>%
  ungroup()

downsampled_df

```

# Look at public vs. private

```{r}

downsampled_public_df_list <- downsampled_df %>%
  group_by(cdr3, v_gene, j_gene, c_gene) %>%
  filter(n_distinct(Patient_ID) > 1) %>%
  mutate(clonotype = paste(cdr3, v_gene, j_gene, c_gene, sep = ""))

downsampled_public_df_list

downsampled_public_df <- downsampled_df %>%
  mutate(clonotype = paste(cdr3, v_gene, j_gene, c_gene, sep = "")) %>%
  filter(clonotype %in% downsampled_public_df_list$clonotype,
         predicted.celltype.l1 == "CD4 T",
         predicted.celltype.l2 == "CD4 TCM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  mutate(tag = "Public\n(downsampled)")

downsampled_public_df

public_df_list <- combined_data %>%
  group_by(cdr3, v_gene, j_gene, c_gene) %>%
  filter(n_distinct(Patient_ID) > 1) %>%
  mutate(clonotype = paste(cdr3, v_gene, j_gene, c_gene, sep = ""))

public_df_list

public_df <- combined_data %>%
  mutate(clonotype = paste(cdr3, v_gene, j_gene, c_gene, sep = "")) %>%
  filter(clonotype %in% public_df_list$clonotype,
         predicted.celltype.l1 == "CD4 T",
         predicted.celltype.l2 == "CD4 TCM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  mutate(tag = "Public")

public_df

```

# Look at TCR convergence

```{r}

convergence_within_patients <- combined_data %>%
  select(Patient_ID, barcode, Chain, cdr3, cdr3_nt, v_gene, d_gene, j_gene, c_gene) %>%
  group_by(cdr3, Patient_ID, v_gene) %>%
  filter(n_distinct(cdr3_nt) > 1) %>%
  group_by(Patient_ID, cdr3, cdr3_nt) %>%
  summarize(count = n())

convergence_across_patients <- combined_data %>%
  mutate(clonotype = paste(cdr3, v_gene, j_gene, c_gene, sep = "")) %>%
  group_by(cdr3, v_gene) %>%
  filter(n_distinct(cdr3_nt) > 1)
  #filter(clonotype %in% public_df_list$clonotype) %>%
  #group_by(Patient_ID, cdr3, cdr3_nt) %>%
  #summarize(count = n())

convergence_across_patients

# do see more convergence across patients as expected, and they are mostly in public TCRs also as expected

convergence_df <- combined_data %>%
  mutate(clonotype = paste(cdr3, v_gene, j_gene, c_gene, sep = "")) %>%
  filter(clonotype %in% convergence_across_patients$clonotype,
         predicted.celltype.l1 == "CD4 T",
         predicted.celltype.l2 == "CD4 TCM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  mutate(tag = "Converged")

convergence_df

```

# Look at overlap between public and converged

```{r}

# Create a Venn diagram
venn.plot <- venn.diagram(
  x = list(A = public_df$barcode, B = convergence_df$barcode),
  category.names = c("Public TCRs", "Converged TCRs"),
  filename = NULL,
  output = TRUE
)

# Display the Venn diagram
grid.draw(venn.plot)

```

# Get unique CD4 TCMs TRBs from top 15 clonotypes per irAE group (normalizing counts by seq depth)

```{r}

unique_cd8_tem_trb <- combined_data %>%
  filter(predicted.celltype.l1 == "CD4 T",
         predicted.celltype.l2 == "CD4 TCM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # normalize by sequencing depth
  left_join(seq_depth, by = c("Patient_ID", "irAE")) %>%
  mutate(normalized_count = count / seq_depth) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  arrange(-normalized_count) %>%
  ungroup() %>%
  mutate(tag = "All")

unique_cd8_tem_trb

unique_cd8_tem_trb_top_15 <- unique_cd8_tem_trb %>%
  group_by(predicted.celltype.l1, Chain, irAE) %>%
  filter(row_number() <= 15) %>%
  ungroup() %>%
  mutate(tag = "Top 15 clonotypes\n(normalized)")

unique_cd8_tem_trb_top_15

```

# Get unique CD8 Naive TRBs from top n clonotypes per irAE group (not normalizing counts by seq depth)

```{r}

unique_cd8_tem_trb_top_5_not_normalized <- combined_data %>%
  filter(predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  arrange(-count) %>%
  ungroup() %>%
  group_by(predicted.celltype.l1, Chain, irAE) %>%
  filter(row_number() <= 5) %>%
  ungroup() %>%
  mutate(tag = "Top 5 clonotypes\n(not normalized)")

unique_cd8_tem_trb_top_10_not_normalized <- combined_data %>%
  filter(predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  arrange(-count) %>%
  ungroup() %>%
  group_by(predicted.celltype.l1, Chain, irAE) %>%
  filter(row_number() <= 10) %>%
  ungroup() %>%
  mutate(tag = "Top 10 clonotypes\n(not normalized)")

unique_cd8_tem_trb_top_15_not_normalized <- combined_data %>%
  filter(predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  arrange(-count) %>%
  ungroup() %>%
  group_by(predicted.celltype.l1, Chain, irAE) %>%
  filter(row_number() <= 15) %>%
  ungroup() %>%
  mutate(tag = "Top 15 clonotypes\n(not normalized)")

unique_cd8_tem_trb_top_20_not_normalized <- combined_data %>%
  filter(predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  arrange(-count) %>%
  ungroup() %>%
  group_by(predicted.celltype.l1, Chain, irAE) %>%
  filter(row_number() <= 20) %>%
  ungroup() %>%
  mutate(tag = "Top 20 clonotypes\n(not normalized)")

unique_cd8_tem_trb_top_25_not_normalized <- combined_data %>%
  filter(predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  arrange(-count) %>%
  ungroup() %>%
  group_by(predicted.celltype.l1, Chain, irAE) %>%
  filter(row_number() <= 25) %>%
  ungroup() %>%
  mutate(tag = "Top 25 clonotypes\n(not normalized)")

```

# Join dfs

```{r}

unique_cd8_tem_trb_all <- unique_cd8_tem_trb_top_20_not_normalized %>%
  full_join(unique_cd8_tem_trb_top_15_not_normalized) %>%
  full_join(unique_cd8_tem_trb_top_10_not_normalized) %>%
  full_join(unique_cd8_tem_trb_top_5_not_normalized)

```

# Plot pgen conclusions

```{r}

unique_cd8_tem_trb_all$tag <- factor(unique_cd8_tem_trb_all$tag, levels = c("Top 5 clonotypes\n(not normalized)", "Top 10 clonotypes\n(not normalized)", "Top 15 clonotypes\n(not normalized)", "Top 20 clonotypes\n(not normalized)"))

p <- ggplot(unique_cd8_tem_trb_all, aes(x = tag, y = log10(pgen), fill = irAE)) +
  geom_boxplot() +
  labs(x = "", fill = "irAE", y = "CD8 TEM TRB\npgen (log10)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

# Adjust y-axis scales with a specified number of ticks
p + scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

# Filter the data for statistical test
test_data <- unique_cd8_tem_trb_all %>%
  filter(tag == "Top 15 clonotypes\n(not normalized)")

# Calculate the Wilcoxon rank-sum test statistic and p-value
wilcox_result <- wilcox.test(test_data$pgen ~ test_data$irAE)

# Print the results
print(wilcox_result)

```

# Plot length conclusions

```{r}

p <- ggplot(unique_cd8_tem_trb_all, aes(x = tag, y = (cdr3_length), fill = irAE)) +
  geom_boxplot() +
  labs(x = "", fill = "irAE", y = "CD4 TCM TRB\nCDR3 length (AAs)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

# Adjust y-axis scales with a specified number of ticks
p + scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

# Filter the data for statistical test
test_data <- unique_cd8_tem_trb_all %>%
  filter(tag == "Top 5 clonotypes\n(not normalized)")

# Calculate the Wilcoxon rank-sum test statistic and p-value
wilcox_result <- wilcox.test(test_data$cdr3_length ~ test_data$irAE)

# Print the results
print(wilcox_result)

```