---
title: "Identification of Pathogenic Immune Cell Subsets Associated with Checkpoint Inhibitor-induced Myocarditis"
output: pdf_document
---

 - data reproduction of https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9397491/
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries

```{r}

library(Seurat)
library(Matrix)
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(circlize)
library(scales)
library(ggpubr)

```

# Set theme

```{r}

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_rect(colour="black", fill = NA, size = 1),
          axis.text = element_text(colour="black"),
          axis.ticks = element_line(colour="black"),
          axis.text.x = element_text(angle=0),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 7)))

```

# Read in final RNA T cell Seurat with TCR metadata

```{r}

tcr_rna <- readRDS("../saved_data/final_seurat_T_cells.rds")

```

# Tibble metadata, add rownames

```{r}

dehashed_combined_tcr_seq <- as_tibble(tcr_rna@meta.data)

dehashed_combined_tcr_seq$barcode <- colnames(tcr_rna)

dehashed_combined_tcr_seq <- dehashed_combined_tcr_seq %>%
  pivot_longer(cols = matches("(TRA|TRB)$"), names_to = c(".value", "Chain"), names_pattern = "(.*)_([A-Z]+)$") %>%
  mutate(time_point = ifelse(grepl("[EL]", Patient_ID), sub(".*([EL]).*", "\\1", Patient_ID), NA), Patient_ID = gsub("[EL]", "", Patient_ID))

dehashed_combined_tcr_seq

write.csv(dehashed_combined_tcr_seq, "../saved_data/dehashed_tcr_seq_with_rna.csv")

```

# Assess cell identity confidence

```{r}

cell_confidence_test <- dehashed_combined_tcr_seq %>%
  select(barcode, chain, cdr3, Patient_ID, irAE, predicted.celltype.l1, predicted.celltype.l1.score, predicted.celltype.l2, predicted.celltype.l2.score) %>%
  group_by(predicted.celltype.l1, predicted.celltype.l2) %>%
  summarize(count = n(), score1 = mean(predicted.celltype.l1.score), score2 = mean(predicted.celltype.l2.score))

cell_confidence_test

# unexpected combos have low scores, remove them

```

# Prep feature df

```{r}

feature_df <- dehashed_combined_tcr_seq %>%
  filter(
    case_when(
      predicted.celltype.l1 == "CD4 T" & str_detect(predicted.celltype.l2, "CD4") | predicted.celltype.l2 == "Treg" ~ TRUE,
      predicted.celltype.l1 == "CD8 T" & str_detect(predicted.celltype.l2, "CD8") ~ TRUE,
      predicted.celltype.l1 == "other T" & predicted.celltype.l2 %in% c("MAIT", "dnT") ~ TRUE,
      TRUE ~ FALSE)) %>%
  select(-cdr3_nt) %>%
  rowwise() %>%
  mutate(cdr3_length = nchar(cdr3)) %>%
  ungroup() %>%
  mutate(group = case_when(
    startsWith(Patient_ID, "B") ~ "non-myocarditis irAE",
    startsWith(Patient_ID, "M") ~ "myocarditis irAE",
    startsWith(Patient_ID, "A") ~ "no irAE",
    TRUE ~ "other"  # Add a default condition if none of the above match
  ))

write.csv(feature_df, "../saved_data/tcr_seq_features.csv")

feature_df

```

# Examine patient sampling depth bias

```{r}

feature_df %>%
  group_by(Patient_ID, irAE) %>%
  summarize(count = n()) %>%
  arrange(-count) %>%
  print()

# Choose 328 seqs from non B10 samples to downsample all to 328

```

# Normalize cell numbers across samples

```{r}

desired_rows_per_patient <- 328/2
# /2 to get equal numbers of each chain (TRA and TRB)

set.seed(123)

# Downsample the dataframe
downsampled_df <- feature_df %>%
  group_by(Patient_ID, Chain) %>%
  sample_n(desired_rows_per_patient, replace = FALSE) %>%
  ungroup()

downsampled_df

downsampled_df_doubled <- downsampled_df %>%
  bind_rows(mutate(downsampled_df, predicted.celltype.l1 = "All T cells"))

downsampled_df_doubled

```

# Look at public vs. private

```{r}

downsampled_public_df <- downsampled_df %>%
  group_by(cdr3, v_gene, j_gene, c_gene) %>%
  filter(n_distinct(Patient_ID) > 1) %>%
  select(Patient_ID, cdr3) %>%
  mutate(clonotype = paste(cdr3, v_gene, j_gene, c_gene, sep = ""))

downsampled_public_df

downsampled_df <- downsampled_df %>%
  mutate(clonotype = paste(cdr3, v_gene, j_gene, c_gene, sep = "")) %>%
  mutate(publicness = if_else(clonotype %in% downsampled_public_df$clonotype, "public", "private"))

downsampled_df

public_df <- feature_df %>%
  group_by(cdr3, v_gene, j_gene, c_gene) %>%
  filter(n_distinct(Patient_ID) > 1) %>%
  select(Patient_ID, cdr3) %>%
  mutate(clonotype = paste(cdr3, v_gene, j_gene, c_gene, sep = ""))

public_df

feature_df <- feature_df %>%
  mutate(clonotype = paste(cdr3, v_gene, j_gene, c_gene, sep = "")) %>%
  mutate(publicness = if_else(clonotype %in% public_df$clonotype, "public", "private"))

feature_df
  
```

# Import pgen data

```{r}

igor_alpha <- read.csv("../saved_data/igor_output_alpha.csv") %>%
  dplyr::rename(TRA_pgen = pgen) %>%
  filter(!(is.na(TRA_pgen)),
         TRA_pgen != "NaN")

igor_combined <- read.csv("../saved_data/igor_output_beta.csv") %>%
  dplyr::rename(TRB_pgen = pgen) %>%
  filter(!(is.na(TRB_pgen)),
         TRB_pgen != "NaN") %>%
  full_join(igor_alpha, by = c("TCR_name", "TRA_nt", "TRB_nt", "TRA_aa", "TRB_aa", "TRAV", "TRAJ", "TRA_CDR3", "TRBV", "TRBJ", "TRB_CDR3", "TRAC", "TRBC", "TRA_leader", "TRB_leader", "Linker", "Link_order", "TRA_5_prime_seq", "TRA_3_prime_seq", "TRB_5_prime_seq", "TRB_3_prime_seq", "Linked_nt", "Linked_aa", "Warnings.Errors")) %>%
  select(TCR_name, TRA_CDR3, TRB_CDR3, TRA_pgen, TRB_pgen) %>%
  mutate(TRA_pgen = as.character(TRA_pgen),
         TRB_pgen = as.character(TRB_pgen))

```

# Get CD8 TEMs TRBs unique dfs

```{r}

unique_downsampled_igor_combined_cd8_tems <- igor_combined %>%
  pivot_longer(cols = matches("^(TRA|TRB)"),
               names_to = c("Chain", ".value"),
               names_pattern = "^(TRA|TRB)_(.*)$") %>%
  dplyr::rename(barcode = TCR_name,
         cdr3 = CDR3) %>%
  mutate(pgen = as.numeric(pgen)) %>%
  right_join(downsampled_df, by = c("barcode", "Chain", "cdr3")) %>%
  filter(predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  mutate(count = n()) %>%
  dplyr::slice(1) %>%
  arrange(-count) %>%
  ungroup() %>%
  mutate(clonotype_order = min_rank(dplyr::desc(count)))

unique_downsampled_igor_combined_cd8_tems

unique_full_igor_combined_cd8_tems <- igor_combined %>%
  pivot_longer(cols = matches("^(TRA|TRB)"),
               names_to = c("Chain", ".value"),
               names_pattern = "^(TRA|TRB)_(.*)$") %>%
  dplyr::rename(barcode = TCR_name,
         cdr3 = CDR3) %>%
  mutate(pgen = as.numeric(pgen)) %>%
  right_join(feature_df, by = c("barcode", "Chain", "cdr3")) %>%
  filter(predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  mutate(count = n()) %>%
  dplyr::slice(1) %>%
  arrange(-count) %>%
  ungroup() %>%
  mutate(clonotype_order = min_rank(dplyr::desc(count)))

unique_full_igor_combined_cd8_tems

```

# Create df of just solid conclusions +/- downsampling

```{r}

unique_downsampled_cd8_tems_top_100 <- unique_downsampled_igor_combined %>%
  filter(clonotype_order <= 100) %>%
  mutate(clonotype_rank = "Top 100")

unique_downsampled_cd8_tems_top_20 <- unique_downsampled_igor_combined %>%
  filter(clonotype_order <= 20) %>%
  mutate(clonotype_rank = "Top 20")

unique_downsampled_cd8_tems_top_50 <- unique_downsampled_igor_combined %>%
  filter(clonotype_order <= 50) %>%
  mutate(clonotype_rank = "Top 50")

unique_downsampled_cd8_tems_all <- unique_downsampled_igor_combined %>%
  mutate(clonotype_rank = "All")

unique_downsampled_cd8_tems <- unique_downsampled_cd8_tems_top_100 %>%
  full_join(unique_downsampled_cd8_tems_top_20) %>%
  full_join(unique_downsampled_cd8_tems_top_50) %>%
  full_join(unique_downsampled_cd8_tems_all)

unique_downsampled_cd8_tems

unique_full_cd8_tems_top_20 <- unique_full_igor_combined_cd8_tems %>%
  filter(clonotype_order <= 20) %>%
  mutate(clonotype_rank = "Top 20")

unique_full_cd8_tems_top_30 <- unique_full_igor_combined_cd8_tems %>%
  filter(clonotype_order <= 30) %>%
  mutate(clonotype_rank = "Top 30")

unique_full_cd8_tems_top_50 <- unique_full_igor_combined_cd8_tems %>%
  filter(clonotype_order <= 50) %>%
  mutate(clonotype_rank = "Top 50")

unique_full_cd8_tems_all <- unique_full_igor_combined_cd8_tems %>%
  mutate(clonotype_rank = "All")

unique_full_cd8_tems <- unique_full_cd8_tems_top_20 %>%
  full_join(unique_full_cd8_tems_top_30) %>%
  full_join(unique_full_cd8_tems_top_50) %>%
  full_join(unique_full_cd8_tems_all)

unique_full_cd8_tems

```

# Plot pgen solid conclusions for downsampling

```{r}

p <- ggplot(unique_downsampled_cd8_tems, aes(x = clonotype_rank, y = log10(pgen), fill = irAE)) +
  geom_boxplot() +
  labs(x = "Clonotype rank", fill = "irAE", y = "TRB pgen (log10)")
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

# Adjust y-axis scales with a specified number of ticks
p + scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

unique_downsampled_cd8_tems %>%
  group_by(clonotype_rank, irAE) %>%
  summarize(count = n()) %>%
  print()

# Filter the data for statistical test
test_data <- unique_downsampled_cd8_tems %>%
  filter(clonotype_rank == "Top 100")

# Calculate the Wilcoxon rank-sum test statistic and p-value
wilcox_result <- wilcox.test(test_data$pgen ~ test_data$irAE)

# Print the results
print(wilcox_result)

# top 5: 0.8
# top 20: 0.4445
# top 50: 0.3988

```

# Plot pgen solid conclusions without downsampling

```{r}

unique_full_cd8_tems$clonotype_rank <- factor(unique_full_cd8_tems$clonotype_rank, levels = c("Top 20", "Top 30", "Top 50", "All"))

p <- ggplot(unique_full_cd8_tems, aes(x = clonotype_rank, y = log10(pgen), fill = irAE)) +
  geom_boxplot() +
  labs(x = "Clonotype rank", fill = "irAE", y = "TRB pgen (log10)")

# Adjust y-axis scales with a specified number of ticks
p + scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

unique_full_cd8_tems %>%
  group_by(clonotype_rank, irAE) %>%
  summarize(count = n()) %>%
  print()

# Filter the data for statistical test
test_data <- unique_full_cd8_tems %>%
  filter(clonotype_rank == "All")

# Calculate the Wilcoxon rank-sum test statistic and p-value
wilcox_result <- wilcox.test(test_data$pgen ~ test_data$irAE)

# Print the results
print(wilcox_result)

# top 20: 0.02002
# top 30: 0.004194
# top 50: 0.03725
# All: 0.1051

# adj: 0.04004000 0.01677600 0.04966667 0.10510000

p_values <- c(0.02002, 0.004194, 0.03725, 0.1051)

# Adjust p-values using the Benjamini-Hochberg method (default method)
adjusted_p_values <- p.adjust(p_values, method = "BH")

# Print the adjusted p-values
print(adjusted_p_values)

```

# Plot solid pgen conclusions without downsampling, patient bias

```{r}

p <- ggplot(unique_full_cd8_tems, aes(x = Patient_ID, y = log10(pgen), fill = irAE)) +
  geom_boxplot() +
  facet_wrap(~clonotype_rank) +
    labs(x = "Patient ID", fill = "irAE", y = "TRB pgen (log10)") +
  theme(axis.text.x = element_blank())

# Adjust y-axis scales with a specified number of ticks
p + scale_y_continuous(breaks = pretty_breaks(n = 3))

```

# Plot solid length conclusions for downsampling

```{r}

p <- ggplot(unique_downsampled_cd8_tems, aes(x = clonotype_rank, y = cdr3_length, fill = irAE)) +
  geom_boxplot() +
  labs(x = "", fill = "irAE", y = "CD8 TEM CDR3β\nlength (AAs)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

# Adjust y-axis scales with a specified number of ticks
p + scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

unique_downsampled_cd8_tems %>%
  group_by(clonotype_rank, irAE) %>%
  summarize(count = n()) %>%
  print()

# Filter the data for statistical test
test_data <- unique_downsampled_cd8_tems %>%
  filter(clonotype_rank == "Top 100")

# top 20: 0.2888
# top 100: 0.2735

#adj: 

# Calculate the Wilcoxon rank-sum test statistic and p-value
wilcox_result <- wilcox.test(test_data$cdr3_length ~ test_data$irAE)

# Print the results
print(wilcox_result)

```

# Plot solid length conclusions without downsampling

```{r}

p <- ggplot(unique_full_cd8_tems %>% filter(clonotype_rank %in% c("Top 20", "Top 30", "All")), aes(x = clonotype_rank, y = cdr3_length, fill = irAE)) +
  geom_boxplot() +
  labs(x = "Clonotype rank", fill = "irAE", y = "CD8 TEM CDR3β\nlength (AAs)")

# Adjust y-axis scales with a specified number of ticks
p + scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

unique_full_cd8_tems %>%
  group_by(irAE, clonotype_rank) %>%
  summarize(count = n()) %>%
  print()

# Filter the data for statistical test
test_data <- unique_full_cd8_tems %>%
  filter(clonotype_rank == "All")

# top 20: 0.05834
# top 30: 0.01727
# All: 0.9734

# adj: 0.05834 0.03454

# Calculate the Wilcoxon rank-sum test statistic and p-value
wilcox_result <- wilcox.test(test_data$cdr3_length ~ test_data$irAE)

# Print the results
print(wilcox_result)

p_values <- c(0.05834, 0.01727, 0.9734)

# Adjust p-values using the Benjamini-Hochberg method (default method)
adjusted_p_values <- p.adjust(p_values, method = "BH")

# Print the adjusted p-values
print(adjusted_p_values)

```

# Plot solid length conclusions without downsampling, patient bias

```{r}

p <- ggplot(unique_full_cd8_tems %>% filter(clonotype_rank %in% c("Top 20", "Top 30", "All")), aes(x = Patient_ID, y = cdr3_length, fill = irAE)) +
  geom_boxplot() +
  facet_wrap(~clonotype_rank) +
  labs(x = "Patient ID", fill = "irAE", y = "CD8 TEM CDR3β\nlength (AAs)") +
  theme(axis.text.x = element_blank())

# Adjust y-axis scales with a specified number of ticks
p + scale_y_continuous(breaks = pretty_breaks(n = 3))

```

# See if germlineness and CDR3 length anticorrelate

```{r}

full_igor_combined <- igor_combined %>%
  pivot_longer(cols = matches("^(TRA|TRB)"),
               names_to = c("Chain", ".value"),
               names_pattern = "^(TRA|TRB)_(.*)$") %>%
  dplyr::rename(barcode = TCR_name,
         cdr3 = CDR3) %>%
  mutate(pgen = as.numeric(pgen)) %>%
  right_join(feature_df, by = c("barcode", "Chain", "cdr3")) %>%
  group_by(Patient_ID, Chain, cdr3, v_gene, j_gene) %>%
  dplyr::slice(1)

full_igor_combined

ggplot(aes(x = cdr3_length, y = log10(pgen)), data = full_igor_combined) +
  geom_point() +
  labs(y = "log10(pgen)", x = "CDR3 length (AAs)") +
  stat_cor(method = "pearson", label.x = 0.8, label.y = 0.9)

```