---
title: "Identification of Pathogenic Immune Cell Subsets Associated with Checkpoint Inhibitor-induced Myocarditis"
output: pdf_document
---

 - data mining of https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9397491/
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries

```{r}

library(ggplot2)
library(dplyr)
library(tidyr)
library(VennDiagram)
library(scales)
library(ggpubr)
library(ComplexHeatmap)
library(circlize)

```

# Set theme

```{r}

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_rect(colour="black", fill = NA, size = 1),
          axis.text = element_text(colour="black"),
          axis.ticks = element_line(colour="black"),
          axis.text.x = element_text(angle=0),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 7)))

```

# Read in data

```{r}

feature_df <- read.csv("../saved_data/tcr_seq_features.csv")

feature_df

```

# Prep hydrophobicity calculator

```{r}

# Define the amino acids and their corresponding Eisenberg hydrophobicity values
amino_acids <- c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
hydrophobicity_values <- c(0.620, -2.530, -0.780, -0.900, 0.290, -0.850, -0.740, 0.480, -0.400, 1.380, 1.060, -1.500, 0.640, 1.190, 0.120, -0.180, -0.050, 0.81, 0.26, 1.08)

# Create a data frame
hydrophobicity_df <- data.frame(Amino_Acid = amino_acids, Hydrophobicity = hydrophobicity_values)

hydrophobicity_df

calculate_hydrophobicity <- function(sequence, hydrophobicity_df) {
  sequence <- toupper(sequence)
  # Calculate the hydrophobicity scores for each amino acid, and sum them
  scores <- sapply(strsplit(sequence, '')[[1]], function(aa) {
    match_aa <- match(aa, hydrophobicity_df$Amino_Acid)
    if (!is.na(match_aa)) {
      return(hydrophobicity_df$Hydrophobicity[match_aa])
    } else {
      # If an amino acid is not found, return NA
      return(NA)
    }
  })
  # If there are any NAs in the scores, return NA; otherwise, return the sum
  if (any(is.na(scores))) {
    return(NA)
  } else {
    return(sum(scores))
  }
}

```

# Import pgen data, join with other data

```{r}

igor_alpha <- read.csv("../saved_data/igor_output_alpha.csv") %>%
  dplyr::rename(TRA_pgen = pgen) %>%
  filter(!(is.na(TRA_pgen)),
         TRA_pgen != "NaN")

combined_data <- read.csv("../saved_data/igor_output_beta.csv") %>%
  dplyr::rename(TRB_pgen = pgen) %>%
  filter(!(is.na(TRB_pgen)),
         TRB_pgen != "NaN") %>%
  full_join(igor_alpha, by = c("TCR_name", "TRA_nt", "TRB_nt", "TRA_aa", "TRB_aa", "TRAV", "TRAJ", "TRA_CDR3", "TRBV", "TRBJ", "TRB_CDR3", "TRAC", "TRBC", "TRA_leader", "TRB_leader", "Linker", "Link_order", "TRA_5_prime_seq", "TRA_3_prime_seq", "TRB_5_prime_seq", "TRB_3_prime_seq", "Linked_nt", "Linked_aa", "Warnings.Errors")) %>%
  select(TCR_name, TRA_CDR3, TRB_CDR3, TRA_pgen, TRB_pgen) %>%
  mutate(TRA_pgen = as.character(TRA_pgen),
         TRB_pgen = as.character(TRB_pgen)) %>%
  pivot_longer(cols = matches("^(TRA|TRB)"),
               names_to = c("Chain", ".value"),
               names_pattern = "^(TRA|TRB)_(.*)$") %>%
  dplyr::rename(barcode = TCR_name,
         cdr3 = CDR3) %>%
  mutate(pgen = as.numeric(pgen)) %>%
  right_join(feature_df, by = c("barcode", "Chain", "cdr3")) %>%
  rowwise() %>%
  mutate(cdr3_hydrophobicity = calculate_hydrophobicity(cdr3, hydrophobicity_df)) %>%
  ungroup()

combined_data

```
  
# Examine patient sampling depth bias

```{r}

seq_depth <- combined_data %>%
  group_by(Patient_ID, irAE) %>%
  summarize(seq_depth = n()) %>%
  arrange(seq_depth) %>%
  print()

seq_depth

# Choose 324 seqs from non B10 samples to downsample all to 324

```

# Downsample

```{r}

desired_rows_per_patient <- 324/2
# /2 to get equal numbers of each chain (TRA and TRB)

set.seed(123)

# Downsample the dataframe
downsampled_df <- combined_data %>%
  group_by(Patient_ID, Chain) %>%
  sample_n(desired_rows_per_patient, replace = FALSE) %>%
  ungroup()

downsampled_df

write.csv(downsampled_df, "../saved_data/downsampled_df.csv")

```

# Look at public vs. private

```{r}

downsampled_public_df_list <- downsampled_df %>%
  group_by(cdr3) %>%
  filter(n_distinct(Patient_ID) > 1) %>%
  mutate(barc_chain = paste(barcode, chain, sep = ""))

downsampled_public_df_list

downsampled_public_df <- downsampled_df %>%
  mutate(barc_chain = paste(barcode, chain, sep = "")) %>%
  filter(barc_chain %in% downsampled_public_df_list$barc_chain,
         predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # look at just unique clonotypes from each patient
  dplyr::slice(1) %>%
  mutate(tag = "Public\n(downsampled)")

downsampled_public_df

public_df_list <- combined_data %>%
  group_by(cdr3) %>%
  filter(n_distinct(Patient_ID) > 1) %>%
  mutate(barc_chain = paste(barcode, chain, sep = ""))

public_df_list

public_df <- combined_data %>%
  mutate(barc_chain = paste(barcode, chain, sep = "")) %>%
  filter(barc_chain %in% public_df_list$barc_chain,
         predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # look at just unique clonotypes from each patient
  dplyr::slice(1) %>%
  mutate(tag = "Public")

public_df

```

# Look at TCR convergence

```{r}

convergence_within_patients <- combined_data %>%
  select(Patient_ID, barcode, Chain, cdr3, cdr3_nt, v_gene, d_gene, j_gene, c_gene) %>%
  group_by(cdr3, Patient_ID) %>%
  filter(n_distinct(cdr3_nt) > 1) %>%
  ungroup() %>%
  group_by(Patient_ID, cdr3, cdr3_nt) %>%
  summarize(count = n())

convergence_across_patients <- combined_data %>%
  mutate(barc_chain = paste(barcode, chain, sep = "")) %>%
  group_by(cdr3) %>%
  mutate(distinct_cdr3_nt_count = n_distinct(cdr3_nt)) %>%
  filter(distinct_cdr3_nt_count > 1)

convergence_across_patients

# do see more convergence across patients as expected, and they are mostly in public TCRs also as expected

convergence_df <- combined_data %>%
  mutate(barc_chain = paste(barcode, chain, sep = "")) %>%
  filter(barc_chain %in% convergence_across_patients$barc_chain,
         predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # look at just unique clonotypes from each patient
  dplyr::slice(1) %>%
  mutate(tag = "Converged")

convergence_df

```

# Plot donors/junction (AA) (x) vs. donors/junction (nt) (y)

```{r}

donor_df <- combined_data %>%
  mutate(barc_chain = paste(barcode, chain, sep = "")) %>%
  mutate(convergence  = if_else(barc_chain %in% convergence_across_patients$barc_chain, "Converged", "Not converged"),
         publicity = if_else(barc_chain %in% public_df_list$barc_chain, "Public", "Private")) %>%
  group_by(cdr3) %>%
  mutate(donors_per_junction_aa = n_distinct(Patient_ID)) %>%
  ungroup() %>%
  group_by(cdr3_nt) %>%
  mutate(donors_per_junction_nt = n_distinct(Patient_ID)) %>%
  ungroup() %>%
  # look at just unique CDR3s from each patient
  group_by(cdr3, cdr3_nt, Patient_ID) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(irAE = paste("irAE:", irAE))

donor_df

ggplot(donor_df, aes(x = donors_per_junction_aa, y = donors_per_junction_nt, color = convergence)) +
  geom_jitter(size = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~irAE, ncol = 4) +
  labs(x = "Donors/junction (AA)", y = "Donors/junction (nt)", color = "Convergence")

ggplot(donor_df, aes(x = donors_per_junction_aa, y = donors_per_junction_nt, color = publicity)) +
  geom_jitter(size = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~irAE, ncol = 4) +
  labs(x = "Donors/junction (AA)", y = "Donors/junction (nt)", color = "Publicity")

```

# Look at overlap between public and converged

```{r}

# Create a Venn diagram
#venn.plot <- venn.diagram(
#  x = list(A = public_df$barcode, B = convergence_df$barcode),
#  category.names = c("Public TCRs", "Converged TCRs"),
#  filename = NULL,
#  output = TRUE
#)

# Display the Venn diagram
#grid.draw(venn.plot)

```

# Get unique CD8 TEMs TRBs from top 15 clonotypes per irAE group (normalizing counts by seq depth)

```{r}

unique_cd8_tem_trb <- combined_data %>%
  filter(predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # normalize by sequencing depth
  left_join(seq_depth, by = c("Patient_ID", "irAE")) %>%
  mutate(normalized_count = count / seq_depth) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  arrange(-normalized_count) %>%
  ungroup() %>%
  mutate(tag = "All")

unique_cd8_tem_trb

unique_cd8_tem_trb_top_15 <- unique_cd8_tem_trb %>%
  group_by(predicted.celltype.l1, Chain, irAE) %>%
  filter(row_number() <= 15) %>%
  ungroup() %>%
  mutate(tag = "Top 15 clonotypes\n(normalized)")

unique_cd8_tem_trb_top_15

```

# Get unique CD8 TEMs TRBs from top 15 clonotypes per irAE group (not normalizing counts by seq depth)

```{r}

unique_cd8_tem_trb_top_15_not_normalized <- combined_data %>%
  filter(predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  arrange(-count) %>%
  ungroup() %>%
  group_by(predicted.celltype.l1, Chain, irAE) %>%
  filter(row_number() <= 15) %>%
  ungroup() %>%
  mutate(tag = "Top 15 clonotypes\n(not normalized)")

unique_cd8_tem_trb_top_15_not_normalized

```

# Get unique CD8 TEMs TRBs from top 15 clonotypes per irAE group from downsampled data

```{r}

unique_cd8_tem_trb_downsampled <- downsampled_df %>%
  filter(predicted.celltype.l1 == "CD8 T",
         predicted.celltype.l2 == "CD8 TEM",
         Chain == "TRB") %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # look at just unique TCRs
  dplyr::slice(1) %>%
  arrange(-count) %>%
  ungroup() %>%
  mutate(tag = "Downsampled")

unique_cd8_tem_trb_downsampled

unique_cd8_tem_trb_top_15_downsampled <- unique_cd8_tem_trb_downsampled %>%
  group_by(predicted.celltype.l1, Chain, irAE) %>%
  filter(row_number() <= 15) %>%
  ungroup() %>%
  mutate(tag = "Top 15 clonotypes\n(downsampled)")

unique_cd8_tem_trb_top_15_downsampled

```

# Join dfs

```{r}

unique_cd8_tem_trb_all <- unique_cd8_tem_trb_top_15 %>%
  full_join(unique_cd8_tem_trb) %>%
  full_join(unique_cd8_tem_trb_top_15_not_normalized) %>%
  full_join(unique_cd8_tem_trb_top_15_downsampled) %>%
  full_join(unique_cd8_tem_trb_downsampled) %>%
  full_join(public_df) %>%
  full_join(downsampled_public_df) %>%
  full_join(convergence_df)

```

# Plot pgen conclusions

```{r}

unique_cd8_tem_trb_all$tag <- factor(unique_cd8_tem_trb_all$tag, levels = c("Top 15 clonotypes\n(normalized)", "Top 15 clonotypes\n(not normalized)", "Top 15 clonotypes\n(downsampled)", "Converged", "Public", "Public\n(downsampled)", "All", "Downsampled"))

p <- ggplot(unique_cd8_tem_trb_all, aes(x = tag, y = log10(pgen), fill = irAE)) +
  geom_boxplot() +
  labs(x = "", fill = "irAE", y = "CD8 TEM\nTRB pgen (log10)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

# Adjust y-axis scales with a specified number of ticks
p + scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

# Filter the data for statistical test
test_data <- unique_cd8_tem_trb_all %>%
  filter(tag == "Converged")

# Calculate the Wilcoxon rank-sum test statistic and p-value
wilcox_result <- wilcox.test(test_data$pgen ~ test_data$irAE)

# Print the results
print(wilcox_result)

# top 15 (normalized): 0.03
# top 15 (not normalized): 0.02
# top 15 (downsampled): 0.06
# public (downsampled): 0.06

p_values <- c(0.03, 0.02, 0.06)

# Adjust p-values using Benjamini-Hochberg procedure
adjusted_p_values <- p.adjust(p_values, method = "BH")

# Display the original and adjusted p-values
data.frame(original_p_values = p_values, adjusted_p_values = adjusted_p_values)

```

# Plot length conclusions

```{r}

p <- ggplot(unique_cd8_tem_trb_all, aes(x = tag, y = (cdr3_length), fill = irAE)) +
  geom_boxplot() +
  labs(x = "", fill = "irAE", y = "CD8 TEM\ TRB\nCDR3 length (AAs)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

# Adjust y-axis scales with a specified number of ticks
p + scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

# Filter the data for statistical test
test_data <- unique_cd8_tem_trb_all %>%
  filter(tag == "Public\n(downsampled)")

# Calculate the Wilcoxon rank-sum test statistic and p-value
wilcox_result <- wilcox.test(test_data$cdr3_length ~ test_data$irAE)

# Print the results
print(wilcox_result)

# top 15 (not normalized): 0.009

```

# Plot germlineness vs. CDR3 length

```{r}

ggplot(aes(x = cdr3_length, y = log10(pgen)), data = combined_data) +
  geom_point() +
  labs(y = "log10(pgen)", x = "CDR3 length (AAs)") +
  stat_cor(method = "pearson", label.x = 0.8, label.y = 0.9)

```

# Plot abundance vs. CDR3 length or pgen

```{r}

abundance_df <- combined_data %>%
  group_by(Patient_ID, cdr3_nt, chain, v_gene, j_gene) %>%
  mutate(count = n()) %>%
  slice(1) %>%
  ungroup() %>%
  select(predicted.celltype.l1, predicted.celltype.l2, count, Patient_ID, irAE, barcode, Chain, cdr3, cdr3_length, pgen) %>%
  mutate(irAE = paste("irAE:", irAE))

abundance_df

# Plot with best-fit liness
ggplot(abundance_df %>% filter(predicted.celltype.l1 == "CD8 T" & Chain == "TRB"),
       aes(x = log10(count), y = cdr3_length)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, col = "blue") +
  facet_wrap(~ irAE, scales = "free_y") +
  labs(x = "CD8 TRB clonotype abundance (log10)", fill = "irAE", y = "CD8 TRB CDR3 length") +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

# Plot with best-fit lines
ggplot(abundance_df %>% filter(predicted.celltype.l1 == "CD8 T" & Chain == "TRB"),
       aes(x = log10(count), y = log10(pgen))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, col = "blue") +
  facet_wrap(~ irAE, scales = "free_y") +
  labs(x = "CD8 TRB clonotype abundance (log10)", fill = "irAE", y = "CD8 TRB pgen (log10)") +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

```

# Plot abundance vs. CDR3 length or pgen just for public TCRs

 - now trends are not what I see for highly expanded

```{r}

abundance_df_public <- combined_data %>%
  group_by(cdr3, v_gene, j_gene, c_gene) %>%
  filter(n_distinct(Patient_ID) > 1) %>%
  ungroup() %>%
  group_by(Patient_ID, cdr3_nt, chain, v_gene, j_gene) %>%
  mutate(count = n()) %>%
  slice(1) %>%
  ungroup() %>%
  select(predicted.celltype.l1, predicted.celltype.l2, count, Patient_ID, irAE, barcode, Chain, cdr3, cdr3_length, pgen) %>%
  mutate(irAE = paste("irAE:", irAE))

abundance_df_public

# Plot with best-fit liness
ggplot(abundance_df_public %>% filter(predicted.celltype.l1 == "CD8 T" & Chain == "TRB"),
       aes(x = (count), y = cdr3_length)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, col = "blue") +
  facet_wrap(~ irAE, scales = "free_y") +
  labs(x = "CD8 TRB clonotype abundance (log10)", fill = "irAE", y = "CD8 TRB CDR3 length") +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

# Plot with best-fit lines
ggplot(abundance_df_public %>% filter(predicted.celltype.l1 == "CD8 T" & Chain == "TRB"),
       aes(x = (count), y = log10(pgen))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, col = "blue") +
  facet_wrap(~ irAE, scales = "free_y") +
  labs(x = "CD8 TRB clonotype abundance (log10)", fill = "irAE", y = "CD8 TRB pgen (log10)") +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

```

# Prep data for summary plots (normalized method)

```{r}

prepped_data <- combined_data %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # normalize by sequencing depth
  left_join(seq_depth, by = c("Patient_ID", "irAE")) %>%
  mutate(normalized_count = count / seq_depth) %>%
  # look at just unique clonotypes (CDR3 AA - VJ) from each patient
  dplyr::slice(1) %>%
  ungroup() %>%
  group_by(predicted.celltype.l1, predicted.celltype.l2, Chain) %>%
  arrange(-normalized_count) %>%
  mutate(chunk_size = ceiling(n() * 0.1),
         chunk = 10*ceiling(row_number() / chunk_size))

prepped_data

```

# Perform wilcox rank sum tests for top n clonotypes in chunks of 10% (normalized method)

```{r}

# Function to perform the test up to a specific cutoff of clonotype abundances
perform_wilcox_test <- function(data, cutoff) {
  result <- data %>%
    filter(chunk <= cutoff) %>%
    group_by(predicted.celltype.l1, predicted.celltype.l2, Chain) %>%    
    summarize(p_val_pgen = {
                  if(length(unique(irAE)) != 2) {
                    NA
                  } else {
                    wilcox.test(pgen ~ irAE)$p.value
                  }
                },
              p_val_cdr3_hydrophobicity = {
                  if(length(unique(irAE)) != 2) {
                    NA
                  } else {
                    wilcox.test(cdr3_hydrophobicity ~ irAE)$p.value
                  }
                },
              p_val_cdr3_length = {
                  if(length(unique(irAE)) != 2) {
                    NA
                  } else {
                    wilcox.test(cdr3_length ~ irAE)$p.value
                  }
                }) %>%
    ungroup()
  
  return(result)
}

# List to store results
results_list <- list()

# Iterate through chunk cutoffs
for (num in 1:10) {
  result <- prepped_data %>%
    perform_wilcox_test(cutoff = num*10)

  results_list[[as.character(num*10)]] <- result
}

# Combine results into a single dataframe
results_df <- bind_rows(results_list, .id = "Chunk_limit")

results_df <- results_df %>%
  pivot_longer(cols = c("p_val_pgen", "p_val_cdr3_length", "p_val_cdr3_hydrophobicity"), names_to = "p_val") %>%
  rename(p_val_category = p_val) %>%
  mutate(p_val_category = gsub("p_val_", "", p_val_category)) %>%
  mutate(category = paste(predicted.celltype.l2, Chain, p_val_category))

results_df

```

# Make heatmap (normalized method)

```{r}

top_results <- results_df %>%
  filter(value < 0.05) %>%
  mutate(category = paste(predicted.celltype.l2, Chain, p_val_category))

filtered_results_df <- results_df %>%
  filter(category %in% top_results$category) %>%
  mutate(log_val = -log10(value))

# basically anything red (p val -log10 > 1.3) is significant (p < 0.05)

filtered_results_df$Chunk_limit <- factor(filtered_results_df$Chunk_limit, levels = c("10", "20", "30", "40", "50", "60", "70", "80", "90", "100"))

# Set the cutoff for outlining tiles
cutoff_value <- 0.05  # Replace with your desired cutoff

# Plotting with outlined tiles
ggplot(filtered_results_df, aes(x = Chunk_limit, y = factor(interaction(predicted.celltype.l2, p_val_category, Chain)), fill = log_val)) +
  geom_tile(size = 0.5, color = ifelse(filtered_results_df$value < cutoff_value, "black", NA)) +
  scale_fill_gradientn(colors = colorRampPalette(c("blue", "white", "red"))(20), name = "Wilcox p val (-log10)\n for yes vs. no irAE") +
  labs(x = "Top clonotypes cutoff (% of repertoire across patients)", y = "") +
  theme_minimal()

```

# Try to make prettier heatmap

```{r}

heatmap_df <- filtered_results_df %>%
  select(-value, -category) %>%
  rename(value = log_val)

# Prepare the data in matrix form
heatmap_data <- reshape2::dcast(heatmap_df, predicted.celltype.l2 + p_val_category + Chain ~ Chunk_limit, value.var = "value")

# Extract value columns
mat <- as.matrix(heatmap_data[, -(1:3)])

# Set row names
rownames(mat) <- paste(heatmap_data$predicted.celltype.l2, heatmap_data$p_val_category, heatmap_data$Chain, sep = "_")

ha = HeatmapAnnotation(
    foo = 1:10, 
    bar = sample(letters[1:3], 10, replace = TRUE),
    col = list(foo = colorRamp2(c(0, 5, 10), c("blue", "white", "red")),
               bar = c("a" = "red", "b" = "green", "c" = "blue")
    )
)

# Plot the heatmap without clustering
Heatmap(mat,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        top_annotation = ha)

```

# Make some box plots for tiles of interest

```{r}

p <- ggplot(prepped_data %>% filter(predicted.celltype.l2 == "CD8 TEM" & Chain == "TRB" & chunk %in% c(10)), aes(x = factor(chunk), y = log10(pgen), fill = irAE)) +
  geom_boxplot() +
  labs(x = "Top clonotypes cutoff (% of repertoire across patients)", y = "CD8 TEM\nTRB pgen (log10)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12))

# Adjust y-axis scales with a specified number of ticks
p + scale_y_continuous(expand = expansion(mult = c(0.1, 0.5)), breaks = pretty_breaks(n = 3))

# Filter the data for statistical test
test_data <- prepped_data %>%
  filter(predicted.celltype.l2 == "CD8 TEM" & Chain == "TRB" & chunk == 10)

# Calculate the Wilcoxon rank-sum test statistic and p-value
wilcox_result <- wilcox.test(test_data$pgen ~ test_data$irAE)

# Print the results
print(wilcox_result)

```

# Prep data for summary plots (downsampled method)

```{r}

prepped_data_downsampled <- downsampled_df %>%
  group_by(Patient_ID, cdr3, chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # look at just unique clonotypes (CDR3 AA - VJ) from each patient
  dplyr::slice(1) %>%
  ungroup() %>%
  group_by(predicted.celltype.l1, predicted.celltype.l2, Chain) %>%
  arrange(-count) %>%
  mutate(chunk_size = ceiling(n() * 0.1),
         chunk = 10*ceiling(row_number() / chunk_size))

prepped_data_downsampled

```

# Perform wilcox rank sum tests for top n clonotypes in chunks of 10% (downsampled method)

```{r}

# List to store results
results_list_downsampled <- list()

# Iterate through chunk cutoffs
for (num in 1:10) {
  result <- prepped_data_downsampled %>%
    perform_wilcox_test(cutoff = num*10)

  results_list[[as.character(num*10)]] <- result
}

# Combine results into a single dataframe
results_df_downsampled <- bind_rows(results_list, .id = "Chunk_limit")

results_df_downsampled <- results_df_downsampled %>%
  pivot_longer(cols = c("p_val_pgen", "p_val_cdr3_length", "p_val_cdr3_hydrophobicity"), names_to = "p_val") %>%
  rename(p_val_category = p_val) %>%
  mutate(p_val_category = gsub("p_val_", "", p_val_category)) %>%
  mutate(category = paste(predicted.celltype.l2, Chain, p_val_category))

results_df_downsampled

```

# Make heatmap (downsampled method)

```{r}

top_results_downsampled <- results_df_downsampled %>%
  filter(value < 0.05) %>%
  mutate(category = paste(predicted.celltype.l2, Chain, p_val_category))

filtered_results_df_downsampled <- results_df_downsampled %>%
  filter(category %in% top_results_downsampled$category)

# basically anything red (p val -log10 > 1.3) is significant (p < 0.05)

filtered_results_df_downsampled$Chunk_limit <- factor(filtered_results_df_downsampled$Chunk_limit, levels = c("10", "20", "30", "40", "50", "60", "70", "80", "90", "100"))

# Set the cutoff for outlining tiles
cutoff_value <- 0.05  # Replace with your desired cutoff

# Plotting with outlined tiles
ggplot(filtered_results_df_downsampled, aes(x = Chunk_limit, y = factor(interaction(predicted.celltype.l2, p_val_category, Chain)), fill = -log10(value))) +
  geom_tile(size = 0.5, color = ifelse(filtered_results_df_downsampled$value < cutoff_value, "black", NA)) +
  scale_fill_gradientn(colors = colorRampPalette(c("blue", "white", "red"))(20), name = "Wilcox p val (-log10)\n for yes vs. no irAE") +
  labs(x = "Top clonotypes cutoff (% of repertoire across patients)", y = "") +
  theme_minimal()

```