---
title: "T cell characteristics associated with irAE from ICI in melanoma"
output: pdf_document
---

 - data reproduction of https://www.nature.com/articles/s41591-021-01623-z

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries

```{r}

library(Seurat)
library(patchwork)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tidyverse)

```

# Set theme

```{r}

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_rect(colour="black", fill = NA, size = 1),
          axis.text = element_text(colour="black"),
          axis.ticks = element_line(colour="black"),
          axis.text.x = element_text(angle=0),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 7)))

```

# Read in data

```{r}

melanoma_T_cells_cell_type_clustered <- readRDS("../saved_data/cell_type_clustered.rds")

```

# Read in TCR, BCR data, process to add in as metadata to Seurat RNA assay object

```{r}

# List of file names
file_names <- c(
  "../raw_data/GSM5694801_YUFURL_TCR.csv.gz",
  "../raw_data/GSM5694811_YUALOE_TCR.csv.gz",
  "../raw_data/GSM5694802_YUGRUS_TCR.csv.gz",
  "../raw_data/GSM5694812_YUHONEY_TCR.csv.gz",
  "../raw_data/GSM5694803_YUHERN_TCR.csv.gz",
  "../raw_data/GSM5694804_YUKEND_TCR.csv.gz",
  "../raw_data/GSM5694805_YUPIXEL_TCR.csv.gz",
  "../raw_data/GSM5694806_YUROD_TCR.csv.gz",
  "../raw_data/GSM5694798_YUCARD_TCR.csv.gz",
  "../raw_data/GSM5694808_YUTHEA_TCR.csv.gz",
  "../raw_data/GSM5694799_YUENZO_TCR.csv.gz",
  "../raw_data/GSM5694800_YUFUB_TCR.csv.gz",
  "../raw_data/GSM5694810_YUVARDO_TCR.csv.gz",
  "../raw_data/GSM5694809_YUTORY_TCR.csv.gz",
  "../raw_data/GSM5694807_YUTAUR_TCR.csv.gz",
  "../raw_data/GSM5694813_YUNANCY_TCR.csv.gz",
  "../raw_data/GSM5694821_YUPIXEL_BCR.csv.gz",
  "../raw_data/GSM5694822_YUROD_BCR.csv.gz",
  "../raw_data/GSM5694823_YUTAUR_BCR.csv.gz",
  "../raw_data/GSM5694814_YUCARD_BCR.csv.gz",
  "../raw_data/GSM5694824_YUTHEA_BCR.csv.gz",
  "../raw_data/GSM5694815_YUENZO_BCR.csv.gz",
  "../raw_data/GSM5694825_YUTORY_BCR.csv.gz",
  "../raw_data/GSM5694816_YUFUB_BCR.csv.gz",
  "../raw_data/GSM5694826_YUVARDO_BCR.csv.gz",
  "../raw_data/GSM5694817_YUFURL_BCR.csv.gz",
  "../raw_data/GSM5694818_YUGRUS_BCR.csv.gz",
  "../raw_data/GSM5694819_YUHERN_BCR.csv.gz",
  "../raw_data/GSM5694820_YUKEND_BCR.csv.gz"
)

# List of patient IDs you want to include
selected_patient_IDs <- c(
  "YUCARD", "YUPIXEL", "YUENZO", "YUKEND", "YUTHEA", 
  "YUFUB", "YUTAUR", "YUVARDO", "YUGRUS", "YUFURL", 
  "YUTORY", "YUHERN", "YUROD"
)

# Filter the list of file names to include only selected patients
selected_files <- file_names[sapply(file_names, function(file_name) {
  patient_ID <- gsub(".*_(Y[A-Z]+)_.*", "\\1", file_name)
  patient_ID %in% selected_patient_IDs
})]

# Create a mapping of patient.ID to their respective suffixes
suffix_mapping <- data.frame(
  patient.ID = c(
    "YUCARD", "YUPIXEL", "YUENZO", "YUKEND", "YUTHEA", 
    "YUFUB", "YUTAUR", "YUVARDO", "YUGRUS", "YUFURL", 
    "YUTORY", "YUHERN", "YUROD"
  ),
  suffix = c(
    ".1_1", ".1_2", ".1_3", ".1_4", ".1_5", ".1_6", ".1_7", ".1_8", ".1_9", ".1_10", ".1_11", ".1_12", ".1_13"
  )
)

# Create an empty data frame to store the combined data
combined_data <- data.frame()

# Process each file and append it to the combined_data data frame
for (file_name in selected_files) {
  # Read the CSV file
  data <- read_csv(file_name)
  
  # Extract patient ID and CR type (TCR or BCR) from the file name
  patient_ID <- gsub(".*_(Y[A-Z]+)_.*", "\\1", file_name)
  
  # Find the suffix based on patient ID from the mapping
  suffix <- suffix_mapping$suffix[suffix_mapping$patient.ID == patient_ID]
  
  # Add the suffix to the "Barcode" column
  data$Barcode <- paste(data$Barcode, suffix, sep = "")
  
  CR_type <- ifelse(grepl("_TCR", file_name), "TCR", "BCR")
  
  # Add patient ID and CR type as new columns
  data$patient.ID <- patient_ID
  data$CR.type <- CR_type
  
  # Append the data to the combined_data data frame
  combined_data <- bind_rows(combined_data, data)
}

# Remove duplicate rows based on the "Barcode" column
combined_data <- combined_data %>%
  distinct(Barcode, .keep_all = TRUE)

# Set row names based on the "Barcode" column
rownames(combined_data) <- combined_data$Barcode

combined_data <- combined_data %>% select(-Barcode)

combined_data

```

# Add TCR/BCR metadata to Seurat scRNA assay

```{r}

rna_and_tcr_bcr_seurat <- AddMetaData(object = melanoma_T_cells_cell_type_clustered, metadata = combined_data)

melanoma_T_cells_cell_type_clustered

# not every barcode from scRNAseq (24,807 rows) has corresponding TCR/BCR metadata (9,003 barcodes), perhaps partially because I removed duplicate TCR/BCR data barcodes that have same clonotype? 17,440 barcodes if I didn't remove duplicates

```

# Assess aberrant TCR/BCR "expression"

```{r}

# subset Seurat for containing TCR/BCR seq data
tcr_subset <- rna_and_tcr_bcr_seurat[, rna_and_tcr_bcr_seurat@meta.data$CR.type %in% c("TCR")]

# create df
tcr_subset_df <- tibble(cellID = colnames(tcr_subset), clusterID = Idents(tcr_subset))

summary_tcr <- tcr_subset_df %>%
  group_by(clusterID) %>%
  summarize(count = n()) %>%
  arrange(count)

print(summary_tcr)

# non-T cells with "TCRs": 1 monocyte, 28 NKs
# T cells with TCRs: 428 CD8, 518 T/NKT, 3935 CD4 T cells

bcr_subset <- rna_and_tcr_bcr_seurat[, rna_and_tcr_bcr_seurat@meta.data$CR.type %in% c("BCR")]

bcr_subset_df <- tibble(cellID = colnames(bcr_subset), clusterID = Idents(bcr_subset))

summary_bcr <- bcr_subset_df %>%
  group_by(clusterID) %>%
  summarize(count = n()) %>%
  arrange(count)

print(summary_bcr)

# non-B cells with "BCRs": 3 DCs, 15 CD4 T cells, 21 monocytes
# B cells with BCRs: 2706

# I'm inferring from their 2.2% effective doublet rate that my numbers here are a bit off... perhaps due to different Seurat version or manual cell cluster assignments despite cell type clustered UMAP looking pretty similar?

```

# Eliminate cells with aberrant "expression" of TCR/BCR clonotypic sequences

```{r}

# Subset Seurat to remove cells aberrantly expressing TCRs
no_aberrant_rna_and_tcr_bcr_seurat <- rna_and_tcr_bcr_seurat[, rna_and_tcr_bcr_seurat@meta.data$CR.type == "TCR" & !(rna_and_tcr_bcr_seurat@active.ident %in% c("CD8 T cell", "CD4 T cell", "T/NKT")), invert = TRUE]

# Subset Seurat to remove cells aberrantly expressing BCRs
no_aberrant_rna_and_tcr_bcr_seurat <- no_aberrant_rna_and_tcr_bcr_seurat[, no_aberrant_rna_and_tcr_bcr_seurat@meta.data$CR.type == "BCR" & !(no_aberrant_rna_and_tcr_bcr_seurat@active.ident %in% c("B cell")), invert = TRUE]

```

# Compare sizes of Seurat before/after TCR/BCR subsetting, confirm we only lose the expected 68 cells

```{r}

length(colnames(rna_and_tcr_bcr_seurat)) - length(colnames(no_aberrant_rna_and_tcr_bcr_seurat))

```

# Repeat dimensional reduction (PCA, UMAP) and FindClusters since filtering out RBCs, platelets, and cells aberrantly "expressing" TCRs/BCRs

```{r}

no_aberrant_rna_and_tcr_bcr_seurat <- RunPCA(no_aberrant_rna_and_tcr_bcr_seurat, npcs = 30, verbose = FALSE)
no_aberrant_rna_and_tcr_bcr_seurat <- RunUMAP(no_aberrant_rna_and_tcr_bcr_seurat, reduction = "pca", dims = 1:30)
no_aberrant_rna_and_tcr_bcr_seurat <- FindNeighbors(no_aberrant_rna_and_tcr_bcr_seurat, reduction = "pca", dims = 1:30)
no_aberrant_rna_and_tcr_bcr_seurat <- FindClusters(no_aberrant_rna_and_tcr_bcr_seurat, resolution = 3)

```

# Visualize UMAP by cluster

```{r}

DimPlot(no_aberrant_rna_and_tcr_bcr_seurat, reduction = "umap", label = TRUE, repel = TRUE)

ggsave("../figures/umap_cluster_after_first_rbc_platelet_filter_and_tcr_bcr_check.svg")

# see 22 clusters, expect 32 clusters

# I assume that the change comes from the RBC/platelet filtering, but I assume their figure 3A was created with UMAP after the filtering NOT with just UMAP before filtering and not showing the RBCs/platelets (which is how I can recreate their figure)

```

# Find cluster biomarkers

```{r}

# find markers for every cluster compared to all remaining cells, report only the positive ones
no_aberrant_rna_and_tcr_bcr_seurat.markers <- FindAllMarkers(no_aberrant_rna_and_tcr_bcr_seurat, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

```

# Look at relevant cluster biomarkers

```{r}

no_aberrant_rna_and_tcr_bcr_seurat.markers %>%
  group_by(cluster) %>%
  filter(gene %in% c("CD3D", "CD3E", "CD8A", "CD8B", "NKG7", "GNLY", "IL7R", "CD14", "FCGR3A", "FCER1A", "MS4A1", "HBB", "PPBP")) %>%
  arrange(cluster)

```

# Plot marker distribution across clusters to identify CD8 T cells

```{r}

# CD3D/CD3Ehi and CD8A/CD8Bhi and NKG7/GNLYlo = CD8 T cells

# high in CD3D: 0, 2, 3, 6, 13-15, 18, 20
# high in CD3E: 0, 2, 3, 5-6, 13-15, 18, 20
# high in CD8A: 13-14
# high in CD8B: 13-14, 18

# high in NKG7: 5, 8, 10, 13-15
# high in GNLY: 5, 8, 10, 13-15

VlnPlot(no_aberrant_rna_and_tcr_bcr_seurat, features = c("GNLY"), pt.size = 0)

no_aberrant_rna_and_tcr_bcr_seurat.markers %>%
  group_by(cluster) %>%
  filter(gene %in% c("CD3D", "CD3E", "CD8A", "CD8B", "NKG7", "GNLY")) %>%
  arrange(cluster)

# 18 only great looking CD8 T cell cluster...?

```

# Plot marker distribution across clusters to identify CD4 T cells

```{r}

VlnPlot(no_aberrant_rna_and_tcr_bcr_seurat, features = c("IL7R"), pt.size = 0)
# CD3D/CD3Ehi non-CD8 T cells with high IL7R expression and low NKG7/GNLY = CD4 T cells

# high in CD3D: 0, 2, 3, 6, 13-15, 18, 20
# high in CD3E: 0, 2, 3, 5-6, 13-15, 18, 20

# high in CD8A: 13-14
# high in CD8B: 13-14, 18

# high in NKG7: 5, 8, 10, 13-15
# high in GNLY: 5, 8, 10, 13-15

# high in IL7R: 0, 2, 3, 6, 10, 13-15, 18, 20

# 0, 2, 3, 6, 20?

```

# Plot marker distribution across clusters to identify NK cells

```{r}

# NKG7/GNLYhi and CD3D/CD3Elo = NK cells

# high in CD3D: 0, 2, 3, 6, 13-15, 18, 20
# high in CD3E: 0, 2, 3, 5-6, 13-15, 18, 20
# high in CD8A: 13-14
# high in CD8B: 13-14, 18

# high in NKG7: 5, 8, 10, 13-15
# high in GNLY: 5, 8, 10, 13-15

# 8 and 10?

no_aberrant_rna_and_tcr_bcr_seurat.markers %>%
  group_by(cluster) %>%
  filter(gene %in% c("NKG7", "GNLY")) %>%
  arrange(cluster)

```

# Plot marker distribution across clusters to identify DCs

```{r}

VlnPlot(no_aberrant_rna_and_tcr_bcr_seurat, features = c("FCER1A"), pt.size = 0)
# FCER1Ahi = dendritic cells (DCs)
# 21

no_aberrant_rna_and_tcr_bcr_seurat.markers %>%
  group_by(cluster) %>%
  filter(gene %in% c("FCER1A")) %>%
  arrange(cluster)

```

# Plot marker distribution across clusters to identify B cells

```{r}

VlnPlot(no_aberrant_rna_and_tcr_bcr_seurat, features = c("MS4A1"), pt.size = 0)
# MS4A1hi = B cells

# 1, 11, 12, 21 high in MS4A1

no_aberrant_rna_and_tcr_bcr_seurat.markers %>%
  group_by(cluster) %>%
  filter(gene %in% c("MS4A1")) %>%
  arrange(cluster)

# 

```

# Plot marker distribution across clusters to identify monocytes

```{r}

VlnPlot(no_aberrant_rna_and_tcr_bcr_seurat, features = c("CD14"), pt.size = 0)
# CD14 or FCGR3Ahi = monocytes

# CD14hi = 7, 9, 16, 17, 19, 21-22
# FCGR3Ahi = 4, 5, 8, 10, 12-16, 19

# 19?

no_aberrant_rna_and_tcr_bcr_seurat.markers %>%
  group_by(cluster) %>%
  filter(gene %in% c("FCGR3A", "CD14")) %>%
  arrange(cluster)

```

# Plot marker distribution across clusters to identify RBCs and platelets

```{r}

VlnPlot(no_aberrant_rna_and_tcr_bcr_seurat, features = c("HBB"), pt.size = 0)
# HBBhi = red blood cells
# none really, maybe 5, 8-10, 22

no_aberrant_rna_and_tcr_bcr_seurat.markers %>%
  group_by(cluster) %>%
  filter(gene %in% c("HBB")) %>%
  arrange(cluster)

# authors found 2 more RBC clusters

VlnPlot(no_aberrant_rna_and_tcr_bcr_seurat, features = c("PPBP"), pt.size = 0)
# PPBPhi = platelets
# none really

no_aberrant_rna_and_tcr_bcr_seurat.markers %>%
  group_by(cluster) %>%
  filter(gene %in% c("PPBP")) %>%
  arrange(cluster)

```

# Plot marker distribution across clusters to identify T/NKT

```{r}

VlnPlot(no_aberrant_rna_and_tcr_bcr_seurat, features = c("IL7R"), pt.size = 0)
# T/NKT: high expression of CD3D/E and GNLY/NKG7 that were not annotated as CD8/CD4 T cells

# high in CD3D: 0, 2, 3, 6, 13-15, 18, 20
# high in CD3E: 0, 2, 3, 5-6, 13-15, 18, 20

# high in NKG7: 5, 8, 10, 13-15
# high in GNLY: 5, 8, 10, 13-15

# 5, 13-15?

```

# Name cell types for clusters

```{r}

#0: CD4 T cell
#1: B cell
#2: CD4 T cell
#3: CD4 T cell
#4: monocyte
#5: RBC/platelet or T/NKT or monocyte? ideally T/NKT
#6: CD4 T cell
#7: monocyte
#8: RBC/platelet or NK or monocyte? ideally NK
#9: RBC/platelet
#10: RBC/platelet or NK or monocyte? ideally NK
#11: B cell
#12: B cell
#13: T/NKT
#14: T/NKT
#15: T/NKT
#16: RBC/platelet
#17: RBC/platelet
#18: CD8 T cell
#19: monocyte
#20: CD4 T cell
#21: DC or B cell or monocyte? ideally DC otherwise no other DC...
#22: RBC/platelet

new.cluster.ids <- c("CD4 T cell",
                     #0
                     
                     "B cell",
                     #1
                     
                     "CD4 T cell",
                     #2
                     
                     "CD4 T cell",
                     #3
                     
                     "Monocyte",
                     #4
                     
                     "T/NKT",
                     #5
                     
                     "CD4 T cell",
                     #6
                     
                     "Monocyte",
                     #7
                     
                     "NK",
                     #8
                     
                     "RBC/platelet",
                     #9
                     
                     "NK",
                     #10
                     
                     "B cell",
                     #11
                     
                     "B cell",
                     #12
                     
                     "T/NKT",
                     #13
                     
                     "T/NKT",
                     #14
                     
                     "T/NKT",
                     #15
                     
                     "RBC/platelet",
                     #16
                     
                     "RBC/platelet",
                     #17
                     
                     "CD8 T cell",
                     #18
                     
                     "Monocyte",
                     #19
                     
                     "CD4 T cell",
                     #20
                     
                     "DC",
                     #21
                     
                     "RBC/platelet"
                     #22
                     )

names(new.cluster.ids) <- levels(no_aberrant_rna_and_tcr_bcr_seurat)
no_aberrant_rna_and_tcr_bcr_seurat <- RenameIdents(no_aberrant_rna_and_tcr_bcr_seurat, new.cluster.ids)

```

# Plot UMAP with cell type clusters

```{r}

DimPlot(no_aberrant_rna_and_tcr_bcr_seurat, reduction = "umap", label = TRUE, pt.size = 0.5)

ggsave("../figures/umap_cell_type_after_first_rbc_platelet_filter_and_tcr_bcr_check.svg")

# 

```

# Re-filter out RBCs/platelets again

```{r}

rbc_filtered_again_immune.combined <- subset(no_aberrant_rna_and_tcr_bcr_seurat, idents = "RBC/platelet", invert = TRUE)

```

# Plot UMAP with cell clusters after excluding RBC/platelet clusters for the 2nd time

```{r}

DimPlot(rbc_filtered_again_immune.combined, reduction = "umap", label = TRUE, pt.size = 0.1, cols = c('CD4 T cell' = 'red', 'B cell' = 'brown', 'Monocyte' = 'blue', 'T/NKT' = 'darkcyan', 'NK' = 'purple', 'CD8 T cell' = 'green', 'DC' = 'orange'), repel = TRUE, shuffle = TRUE, seed = 992)

ggsave("../figures/rbc_filtered_twice_umap_cell_type.svg")

```

# Repeat dimensional reduction (PCA, UMAP) and FindClusters since filtering out RBCs, platelets for the 2nd time

```{r}

rbc_filtered_again_immune.combined <- RunPCA(rbc_filtered_again_immune.combined, npcs = 30, verbose = FALSE)
rbc_filtered_again_immune.combined <- RunUMAP(rbc_filtered_again_immune.combined, reduction = "pca", dims = 1:30)
rbc_filtered_again_immune.combined <- FindNeighbors(rbc_filtered_again_immune.combined, reduction = "pca", dims = 1:30)
rbc_filtered_again_immune.combined <- FindClusters(rbc_filtered_again_immune.combined, resolution = 3)

```

# Visualize UMAP by cluster

```{r}

DimPlot(rbc_filtered_again_immune.combined, reduction = "umap", label = TRUE, repel = TRUE)

ggsave("../figures/umap_cluster_after_second_rbc_platelet_filter_and_tcr_bcr_check.svg")

# see 25 clusters, expect 32 clusters

# 

```

# Find cluster biomarkers

```{r}

# find markers for every cluster compared to all remaining cells, report only the positive ones
rbc_filtered_again_immune.combined.markers <- FindAllMarkers(rbc_filtered_again_immune.combined, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

```

# Analyze CD4 T cell subclusters

```{r}

# CD4 T cell state 5, lacks expression of CCR7 and SELL (CD62L), consistent with CD4 TEM, see association with severe iRAE development, so I'll need to find this same subcluster they find (combine with subcluster 3)

# maybe start with DEGs across subtypes of CD4 T cells, should see 2 of them closely match each other (their subclusters 3 and 5)

#rbc_filtered_immune.combined %>%
#    group_by(cluster) %>%
#    top_n(n = 10, wt = avg_log2FC) -> top10
#DoHeatmap(pbmc, features = top10$gene) + NoLegend()

```