---
title: "Lineage tracing reveals clonal progenitors and long-term persistence of tumor-specific T cells during immune checkpoint blockade"
output: pdf_document
---

 - data reproduction of https://www.sciencedirect.com/science/article/pii/S153561082300082X
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries

```{r}

library(Seurat)
library(ggplot2)
library(dplyr)
library(readr)
library(readxl)
library(Matrix)

```

# Set theme

```{r}

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_rect(colour="black", fill = NA, size = 1),
          axis.text = element_text(colour="black"),
          axis.ticks = element_line(colour="black"),
          axis.text.x = element_text(angle=0),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 7)))

```

# Read in data

```{r}

# List of sample names
sample_names <- c(
  "GSM5608032_MSK1263_A1",
  "GSM5608033_MSK1263_A3",
  "GSM5608034_MSK1263_A4",
  "GSM5608035_MSK1263_LN",
  "GSM5608036_MSK1263_Normal",
  "GSM5608037_MSK1263_R1",
  "GSM5608038_MSK1263_R2",
  "GSM5608039_MSK1263_R3",
  "GSM5608040_MSK1263_R4",
  "GSM5608041_MSK1263_R5",
  "GSM5608042_MSK1263_R6",
  "GSM5608043_MSK1263_R7",
  "GSM5608044_MSK1263_R8",
  "GSM5608045_MSK1302_LN",
  "GSM5608046_MSK1302_Normal",
  "GSM5608047_MSK1302_R1",
  "GSM5608048_MSK1302_R2",
  "GSM5608049_MSK1302_R3",
  "GSM5608050_MSK1302_R4",
  "GSM5608051_MSK1302_R5",
  "GSM5608052_MSK1302_R6",
  "GSM5608053_MSK1302_R7",
  "GSM5608054_MSK1302_R8",
  "GSM5608055_MSK1344_Normal",
  "GSM5608056_MSK1344_R1",
  "GSM5608057_MSK1344_R2",
  "GSM5608058_MSK1344_R3",
  "GSM5608059_MSK1344_R4",
  "GSM5608060_MSK1344_R5",
  "GSM5608061_MSK1344_R6",
  "GSM5608062_MSK1344_R7",
  "GSM5608063_MSK1344_R8"
)

# Set your data directory
data_dir <- '../raw_data/rna/'

# Initialize an empty list to store Seurat objects
seurat_objects <- list()

# Loop through the sample names and create Seurat objects for each sample
for (sample_name in sample_names) {
  # Construct file paths
  counts_file <- paste0(data_dir, sample_name, "_RNA_matrix.mtx")
  barcodes_file <- paste0(data_dir, sample_name, "_RNA_barcodes.tsv")
  genes_file <- paste0(data_dir, sample_name, "_RNA_features.tsv")
  
  # Read data files
  counts <- readMM(counts_file)
  barcodes <- read.table(barcodes_file, stringsAsFactors = FALSE)[, 1]
  features <- read.csv(genes_file, stringsAsFactors = FALSE, sep = "\t", header = FALSE)
  rownames(counts) <- make.unique(features[, 2])
  colnames(counts) <- barcodes
  
  # Create Seurat object
  seurat_object <- CreateSeuratObject(counts = counts)
  
  # Store sample name as metadata
  seurat_object$sample_name <- sample_name
  
  # Store Seurat object in the list
  seurat_objects[[sample_name]] <- seurat_object
}

```

# Merge Seurat objects from all samples

```{r}

# Merge the first element with a sublist of the rest excluding the first element
merged_seurats <- merge(seurat_objects[[1]], y = seurat_objects[-1])

```

# Calculate percent mitochondrial, ribosomal, hemoglobin features

```{r}

merged_seurats[["percent_mito"]] <- PercentageFeatureSet(merged_seurats, pattern = "^MT-")
merged_seurats[["percent_ribo"]] <- PercentageFeatureSet(merged_seurats, pattern = "^RP[SL]")
merged_seurats[["percent_hb"]] <- PercentageFeatureSet(merged_seurats, pattern = "HB[^(P)]")

```

# Count number of cells from all samples

```{r}

length(colnames(merged_seurats)) # 217101 cells

```

# Count cells per each of the 32 samples

```{r}

cell_capture <- merged_seurats@meta.data %>%
  group_by(sample_name) %>%
  dplyr::count()

cell_capture %>% arrange(n)

```

# Visualize QC metrics across all patient samples

```{r}

vln_plot <- VlnPlot(merged_seurats, features = c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo"), ncol = 4, pt.size=0)

vln_plot

ggsave(vln_plot, file = "../figures/vln_plot_qc.svg")

```

# Set author QC metric thresholds

```{r}

qcMetricsThresholds <-
  c(
    "nFeature_RNA_min" = 450,
    
    "nCount_RNA_max" = 15000,

    "percent_mito" = 15
  )

```

# Plot distributions of nFeature_RNA across samples with nFeature_RNA_min cutoff of 450

```{r}

nFeature_RNA_QC <- ggplot(merged_seurats@meta.data, mapping = aes(x = nFeature_RNA)) +
  geom_density(mapping = aes(color = sample_name), size = 0.5) +
  scale_x_log10() +
  geom_vline(xintercept = qcMetricsThresholds["nFeature_RNA_min"],
             color = "red", linetype = "dashed") +
  theme(legend.position = "none")
ggsave(nFeature_RNA_QC, file = "../figures/nFeature_RNA.svg")

nFeature_RNA_QC

```

# Plot distributions of nCount_RNA across samples

```{r}

nCount_RNA_QC <- ggplot(merged_seurats@meta.data, mapping = aes(x = nCount_RNA)) +
  geom_density(mapping = aes(color = sample_name), size = 0.5) +
  scale_x_log10() +
  geom_vline(xintercept = qcMetricsThresholds["nCount_RNA_max"],
             color = "red", linetype = "dashed") +
  theme(legend.position = "none")
ggsave(nCount_RNA_QC, file = "../figures/nCount_RNA.svg")

nCount_RNA_QC

```

# Plot distributions of percent mito across samples

```{r}

percent_mt_QC <- ggplot(merged_seurats@meta.data, mapping = aes(x = percent_mito + 1/length(colnames(merged_seurats)))) +
  geom_density(mapping = aes(color = sample_name), size = 0.5) +
  scale_x_log10() +
  geom_vline(xintercept = qcMetricsThresholds["percent_mito"],
             color = "red", linetype = "dashed") +
  xlab("percent_mito (with pseudocount)") +
  theme(legend.position = "none")
ggsave(percent_mt_QC, file = "../figures/percent_mt.svg")

percent_mt_QC

```

# Plot distributions of percent ribo across samples

```{r}

percent_rb_QC <- ggplot(merged_seurats@meta.data, mapping = aes(x = percent_ribo + 1/length(colnames(merged_seurats)))) +
  geom_density(mapping = aes(color = sample_name), size = 0.5) +
  scale_x_log10() +
  xlab("percent_ribo (with pseudocount)") +
  theme(legend.position = "none")
ggsave(percent_rb_QC, file = "../figures/percent_rb.svg")

percent_rb_QC

```

# Plot distributions of percent hemoglobin across samples

```{r}

percent_hb_QC <- ggplot(merged_seurats@meta.data, mapping = aes(x = percent_hb + 1/length(colnames(merged_seurats)))) +
  geom_density(mapping = aes(color = sample_name), size = 0.5) +
  scale_x_log10() +
  xlab("percent_hb (with pseudocount)") +
  theme(legend.position = "none")

ggsave(percent_hb_QC, file = "../figures/percent_hb.svg")

percent_hb_QC

# authors didn't mention cutoff for this

```

#Sequencing data from 31 of 32 regional samples passed initial quality control (Figure S2A, Table S1, STAR Methods). After removing the single region that did not pass QC, 63.5–89.9% of the individual cells per region (Figure S2B, Table S1) passed QC filtering, retaining 162,062 high-quality T cells for downstream analyses.

# Filter based on sample-specific QC metrics, track how many drop out

```{r}

# remove 1 problematic sample (GSM5608056_MSK1344_R1)
merged_seurats_filtered_1 <- subset(merged_seurats, sample_name != "GSM5608056_MSK1344_R1")

cellsKeep <-
  WhichCells(
    merged_seurats_filtered_1, expression = nFeature_RNA >= qcMetricsThresholds["nFeature_RNA_min"]) %>%
  intersect(
    WhichCells(
      merged_seurats_filtered_1, expression = nCount_RNA <= qcMetricsThresholds["nCount_RNA_max"])) %>%
  intersect(
    WhichCells(
      merged_seurats_filtered_1, expression = percent_mito <= qcMetricsThresholds["percent_mito"]))
cellsDrop <-
  setdiff(colnames(merged_seurats_filtered_1), cellsKeep)
length(cellsDrop) # 17,968 cells dropped from analysis

# 11,313 cells dropped due to nFeature_RNA_min threshold
cellsDropped_nFeature <- sum(
  merged_seurats_filtered_1$nFeature_RNA < qcMetricsThresholds["nFeature_RNA_min"]
)

# 1,289 cells dropped due to nCount_RNA_max threshold
cellsDropped_nCount <- sum(
  merged_seurats_filtered_1$nCount_RNA > qcMetricsThresholds["nCount_RNA_max"]
)

# 13,473 cells dropped due to percent_mito threshold
cellsDropped_percentMito <- sum(
  merged_seurats_filtered_1$percent_mito > qcMetricsThresholds["percent_mito"]
)

merged_seurats_filtered_2 <-
  subset(merged_seurats_filtered_1, cells = cellsKeep)

length(colnames(merged_seurats_filtered_2)) # 187,077 cells left behind in filtered data

```

# Count cells by sample after QC filtering

```{r}

cell_capture <- merged_seurats_filtered_2@meta.data %>%
  group_by(sample_name) %>%
  dplyr::count()

cell_capture %>% arrange(n)

```

# Save the resulting QC filtered Seurat object

```{r}

saveRDS(merged_seurats_filtered_2, file="../saved_data/merged_seurats_filtered.rds")

```