---
title: "Molecular Pathways of Colon Inflammation Induced by Cancer Immunotherapy"
output: pdf_document
---

 - data mining of https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7415717/
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries

```{r}

library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(scales)
library(ggpubr)
library(ComplexHeatmap)
library(e1071)

```

# Set theme

```{r}

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_rect(colour="black", fill = NA, size = 1),
          axis.text = element_text(colour="black"),
          axis.ticks = element_line(colour="black"),
          axis.text.x = element_text(angle=0),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 7)))

```

# Read in final RNA T cell Seurat with TCR metadata

```{r}

tcr_rna <- read.csv("../saved_data/rna_tcr.csv")

tcr_rna

```

# Assess cell identity confidence

```{r}

cell_confidence_test <- tcr_rna %>%
  group_by(predicted.celltype.l1, predicted.celltype.l2) %>%
  summarize(count = n(), score1 = mean(predicted.celltype.l1.score), score2 = mean(predicted.celltype.l2.score))

cell_confidence_test

# unexpected combos have low scores, remove them

```

# Read in IGoR pgen data

```{r}

pgen_alpha <- read.csv("../saved_data/igor_output_alpha.csv") %>%
  select(TCR_name, pgen) %>%
  mutate(Chain = "TRA")

pgen <- read.csv("../saved_data/igor_output_beta.csv") %>%
  select(TCR_name, pgen) %>%
  mutate(Chain = "TRB") %>%
  full_join(pgen_alpha, by = c("TCR_name", "Chain", "pgen")) %>%
  rename(barcode = TCR_name)

pgen

```

# Prep hydrophobicity calculator

```{r}

# Define the amino acids and their corresponding Eisenberg hydrophobicity values
amino_acids <- c("A", "R", "N", "D", "C", "Q", "E", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
hydrophobicity_values <- c(0.620, -2.530, -0.780, -0.900, 0.290, -0.850, -0.740, 0.480, -0.400, 1.380, 1.060, -1.500, 0.640, 1.190, 0.120, -0.180, -0.050, 0.81, 0.26, 1.08)

# Create a data frame
hydrophobicity_df <- data.frame(Amino_Acid = amino_acids, Hydrophobicity = hydrophobicity_values)

hydrophobicity_df

calculate_hydrophobicity <- function(sequence, hydrophobicity_df) {
  sequence <- toupper(sequence)
  # Calculate the hydrophobicity scores for each amino acid, and sum them
  scores <- sapply(strsplit(sequence, '')[[1]], function(aa) {
    match_aa <- match(aa, hydrophobicity_df$Amino_Acid)
    if (!is.na(match_aa)) {
      return(hydrophobicity_df$Hydrophobicity[match_aa])
    } else {
      # If an amino acid is not found, return NA
      return(NA)
    }
  })
  # If there are any NAs in the scores, return NA; otherwise, return the sum
  if (any(is.na(scores))) {
    return(NA)
  } else {
    return(sum(scores))
  }
}

```

# Prep feature df (update cell types with Trms), join with IGoR pgen data

```{r}

combined_data <- tcr_rna %>%
  mutate(predicted.celltype.l2 = if_else(seurat_clusters %in% c(25, 27), "CD4 Trm", if_else(seurat_clusters %in% c(3, 7, 9, 15, 16), "CD8 Trm", predicted.celltype.l2))) %>%
  filter(
    case_when(
      predicted.celltype.l1 == "CD4 T" & str_detect(predicted.celltype.l2, "CD4") | predicted.celltype.l2 == "Treg" ~ TRUE,
      predicted.celltype.l1 == "CD8 T" & str_detect(predicted.celltype.l2, "CD8") ~ TRUE,
      predicted.celltype.l1 == "other T" & predicted.celltype.l2 %in% c("MAIT", "dnT") ~ TRUE,
      TRUE ~ FALSE)) %>%
  filter(!(is.na(v_gene)),
         !(is.na(d_gene)),
         !(is.na(j_gene)),
         !(is.na(c_gene)),
         !(is.na(cdr3))) %>%
  mutate(group = case_when(
    startsWith(patient, "CT") ~ "no ICI",
    startsWith(patient, "C") ~ "ICI colitis",
    startsWith(patient, "NC") ~ "ICI no colitis")) %>%
  mutate(cdr3_length = nchar(cdr3)) %>%
  ungroup() %>%
  select(-X) %>%
  left_join(pgen, by = c("barcode", "Chain")) %>%
  rowwise() %>%
  mutate(cdr3_hydrophobicity = calculate_hydrophobicity(cdr3, hydrophobicity_df)) %>%
  ungroup()

write.csv(combined_data, "../saved_data/rna_tcr_no_na_good_cell_typing.csv")

combined_data

```

# Examine patient sampling depth bias

```{r}

seq_depth <- combined_data %>%
  group_by(patient, group) %>%
  summarize(seq_depth = n()) %>%
  arrange(seq_depth) %>%
  print()

seq_depth

# Choose ~2512/2 seqs from non B10 samples to downsample all to 2449/2

```

# Downsample

```{r}

desired_rows_per_patient <- 1218
# ~/2 to get equal numbers of each Chain (TRA and TRB)

set.seed(4562)

# Downsample the dataframe
downsampled_df <- combined_data %>%
  group_by(patient, Chain) %>%
  sample_n(desired_rows_per_patient, replace = FALSE) %>%
  ungroup()

downsampled_df

write.csv(downsampled_df, "../saved_data/downsampled_df.csv")

```

# Prep data for summary plots (normalized method)

```{r}

prepped_data <- combined_data %>%
  filter(group != "no ICI") %>%
  group_by(patient, cdr3, Chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # normalize by sequencing depth
  left_join(seq_depth, by = c("patient", "group")) %>%
  mutate(normalized_count = count / seq_depth) %>%
  # look at just unique clonotypes from each patient
  dplyr::slice(1) %>%
  ungroup() %>%
  group_by(predicted.celltype.l1, predicted.celltype.l2, Chain) %>%
  arrange(-normalized_count) %>%
  mutate(chunk_size = ceiling(n() * 0.1),
         chunk = 10*ceiling(row_number() / chunk_size))

prepped_data

```

# Perform wilcox rank sum tests within chunks of 10% clonotypes (normalized method)

```{r}

# Function to perform the test within a specific bin of clonotype abundances
perform_wilcox_test <- function(data, cutoff) {
  result <- data %>%
    filter(chunk == cutoff) %>%
    group_by(predicted.celltype.l1, predicted.celltype.l2, Chain) %>%    
    summarize(
      p_val_pgen = {
        if(length(unique(group)) != 2 | any(table(group) == 1)) {
          NA
        } else {
          wilcox.test(pgen ~ group, exact = FALSE)$p.value
        }
      },
      p_val_cdr3_hydrophobicity = {
        if(length(unique(group)) != 2 | any(table(group) == 1)) {
          NA
        } else {
          wilcox.test(cdr3_hydrophobicity ~ group, exact = FALSE)$p.value
        }
      },
      p_val_cdr3_length = {
        if(length(unique(group)) != 2 | any(table(group) == 1)) {
          NA
        } else {
          wilcox.test(cdr3_length ~ group, exact = FALSE)$p.value
        }
      }
    ) %>%
    ungroup()
  
  return(result)
}

# List to store results
results_list <- list()

# Iterate through chunk cutoffs
for (num in 1:10) {
  result <- prepped_data %>%
    perform_wilcox_test(cutoff = num*10)

  results_list[[as.character(num*10)]] <- result
}

# Combine results into a single dataframe
results_df <- bind_rows(results_list, .id = "Chunk")

results_df <- results_df %>%
  pivot_longer(cols = c("p_val_pgen", "p_val_cdr3_length", "p_val_cdr3_hydrophobicity"), names_to = "p_val") %>%
  rename(p_val_category = p_val) %>%
  mutate(p_val_category = gsub("p_val_", "", p_val_category)) %>%
  mutate(category = paste(predicted.celltype.l2, Chain, p_val_category)) %>%
  # adjust pvals for multiple comparisons
  mutate(value = p.adjust(value, method = "BH")) %>%
  arrange(value)

results_df

```

# Make heatmap (normalized method)

```{r}

top_results <- results_df %>%
  filter(value < 0.05) %>%
  mutate(category = paste(predicted.celltype.l2, Chain, p_val_category))

filtered_results_df <- results_df %>%
  filter(category %in% top_results$category) %>%
  mutate(log_val = -log10(value))

filtered_results_df$Chunk <- factor(filtered_results_df$Chunk, levels = c("10", "20", "30", "40", "50", "60", "70", "80", "90", "100"))

heatmap_df <- filtered_results_df %>%
  select(-value, -category) %>%
  rename(value = log_val) %>%
  mutate(p_val_category = case_when(
    p_val_category == "pgen" ~ "Probability of generation",
    p_val_category == "cdr3_length" ~ "CDR3 length (AAs)",
    p_val_category == "cdr3_hydrophobicity" ~ "CDR3 hydrophobicity"
  )) %>%
  mutate(predicted.celltype.l2 = gsub("CD4 |CD8 ", "", predicted.celltype.l2))

heatmap_df

# Prepare the data in matrix form
heatmap_data <- reshape2::dcast(heatmap_df, predicted.celltype.l1 + predicted.celltype.l2 + p_val_category + Chain ~ Chunk, value.var = "value")

# Extract value columns
mat <- as.matrix(heatmap_data[, -(1:4)])

# Set row names
rownames(mat) <- paste(heatmap_data$predicted.celltype.l1, heatmap_data$predicted.celltype.l2, heatmap_data$p_val_category, heatmap_data$Chain, sep = "_")

# Create a matrix of annotations
chain_annotations <- cbind(
  Chain = heatmap_data$Chain
)

# Create a matrix of annotations
type_annotations <- cbind(
  "Cell type" = heatmap_data$predicted.celltype.l1
)

# Create a matrix of annotations
subtype_annotations <- cbind(
  "Cell subtype" = heatmap_data$predicted.celltype.l2
)

# Create a matrix of annotations
feature_annotations <- cbind(
  Feature = heatmap_data$p_val_category
)

# Plot the heatmap without clustering
Heatmap(mat,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_names_gp = gpar(fontsize = 8),  # Adjust font size if needed
        right_annotation = rowAnnotation("Cell type" = type_annotations,
                                         "Cell subtype" = subtype_annotations,
                                         Chain = chain_annotations,
                                         Feature = feature_annotations,
                                         col = list("Cell type" = c("CD4 T" = "#fde725", "CD8 T" = "#21918c", "other T" = "#440154"),
                                                    "Chain" = c("TRA" = "#fde725", "TRB" = "#440154"),
                                                    "Cell subtype" = c("dnT" = "#440154", "Treg" = "#46327e", "Proliferating" = "#1fa187", "TCM" = "#277f8e",   "TEM" = "#365c8d", "Naive" = "#4ac16d", "MAIT" = "#a0da39", "CTL" = "#fde725", "Trm" = "black"),
                                                    "Feature" = c("CDR3 length (AAs)" = "#21918c", "CDR3 hydrophobicity" = "#fde725", "Probability of generation" = "#440154"))),
        show_row_names = FALSE,
        column_title = "Clonotypes bins\n(% of repertoire across patients)",
        column_title_side = "bottom",
        heatmap_legend_param = list(title = "Wilcox padj val (-log10) for\nyes vs. no colitis irAE"),
        layer_fun = function(j, i, x, y, width, height, fill) {
          v = pindex(mat, i, j)
          l = v > -log10(0.05) & v < -log10(0.01)
          if (any(l)) {
            grid.text("*", x[l], y[l], gp = gpar(fontsize = 10))
          }
          q = v > -log10(0.01)
          if (any(q)) {
            grid.text("**", x[q], y[q], gp = gpar(fontsize = 10))
          }
        }
      )

```

# Prep data for summary plots (downsampled method)

```{r}

prepped_data_downsampled <- downsampled_df %>%
  filter(group != "no ICI") %>%
  group_by(patient, cdr3, Chain, v_gene, j_gene) %>%
  # calculate clonotype counts within patients to get a sense of expansion
  mutate(count = n()) %>%
  # look at just unique clonotypes from each patient
  dplyr::slice(1) %>%
  ungroup() %>%
  group_by(predicted.celltype.l1, predicted.celltype.l2, Chain) %>%
  arrange(-count) %>%
  mutate(chunk_size = ceiling(n() * 0.1),
         chunk = 10*ceiling(row_number() / chunk_size))

prepped_data_downsampled

```

# Perform wilcox rank sum tests within chunks of 10% clonotypes (downsampled method)

```{r}

# List to store results
results_list_downsampled <- list()

# Iterate through chunk cutoffs
for (num in 1:4) {
  result <- prepped_data_downsampled %>%
    perform_wilcox_test(cutoff = num*25)

  results_list[[as.character(num*25)]] <- result
}

# Combine results into a single dataframe
results_df_downsampled <- bind_rows(results_list, .id = "Chunk")

results_df_downsampled <- results_df_downsampled %>%
  pivot_longer(cols = c("p_val_pgen", "p_val_cdr3_length", "p_val_cdr3_hydrophobicity"), names_to = "p_val") %>%
  rename(p_val_category = p_val) %>%
  mutate(p_val_category = gsub("p_val_", "", p_val_category)) %>%
  mutate(category = paste(predicted.celltype.l2, Chain, p_val_category)) %>%
  # adjust pvals for multiple comparisons
  mutate(value = p.adjust(value, method = "BH")) %>%
  arrange(value)

results_df_downsampled

```

# Make heatmap (downsampled method)

```{r}

top_results_downsampled <- results_df_downsampled %>%
  filter(value < 0.05) %>%
  mutate(category = paste(predicted.celltype.l2, Chain, p_val_category))

filtered_results_df_downsampled <- results_df_downsampled %>%
  filter(category %in% top_results_downsampled$category) %>%
  mutate(log_val = -log10(value))

filtered_results_df_downsampled$Chunk <- factor(filtered_results_df_downsampled$Chunk, levels = c("10", "20", "30", "40", "50", "60", "70", "80", "90", "100"))

heatmap_df_downsampled <- filtered_results_df_downsampled %>%
  select(-value, -category) %>%
  rename(value = log_val) %>%
  mutate(p_val_category = case_when(
    p_val_category == "pgen" ~ "Probability of generation",
    p_val_category == "cdr3_length" ~ "CDR3 length (AAs)",
    p_val_category == "cdr3_hydrophobicity" ~ "CDR3 hydrophobicity"
  )) %>%
  mutate(predicted.celltype.l2 = gsub("CD4 |CD8 ", "", predicted.celltype.l2))

heatmap_df_downsampled

# Prepare the data in matrix form
heatmap_data_downsampled <- reshape2::dcast(heatmap_df_downsampled, predicted.celltype.l1 + predicted.celltype.l2 + p_val_category + Chain ~ Chunk, value.var = "value")

# Extract value columns
mat_downsampled <- as.matrix(heatmap_data_downsampled[, -(1:4)])

# Set row names
rownames(mat_downsampled) <- paste(heatmap_data_downsampled$predicted.celltype.l1, heatmap_data_downsampled$predicted.celltype.l2, heatmap_data_downsampled$p_val_category, heatmap_data_downsampled$Chain, sep = "_")

# Create a matrix of annotations
chain_annotations_downsampled <- cbind(
  Chain = heatmap_data_downsampled$Chain
)

# Create a matrix of annotations
type_annotations_downsampled <- cbind(
  "Cell type" = heatmap_data_downsampled$predicted.celltype.l1
)

# Create a matrix of annotations
subtype_annotations_downsampled <- cbind(
  "Cell subtype" = heatmap_data_downsampled$predicted.celltype.l2
)

# Create a matrix of annotations
feature_annotations_downsampled <- cbind(
  Feature = heatmap_data_downsampled$p_val_category
)

# highlight rows of interest
#which_row_highlight = which(grepl("CD8 T_TEM_Probability of generation_TRB", rownames(mat_downsampled)))

#split = data.frame(x = c(rep("A", which_row_highlight - 1), "B", rep("C", nrow(mat_downsampled) - which_row_highlight)))

# Plot the heatmap without clustering
Heatmap(mat_downsampled,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_names_gp = gpar(fontsize = 8),  # Adjust font size if needed
        right_annotation = rowAnnotation("Cell type" = type_annotations_downsampled,
                                         "Cell subtype" = subtype_annotations_downsampled,
                                         Chain = chain_annotations_downsampled,
                                         Feature = feature_annotations_downsampled,
                                         col = list("Cell type" = c("CD4 T" = "#fde725", "CD8 T" = "#21918c", "other T" = "#440154"),
                                                    "Chain" = c("TRA" = "#fde725", "TRB" = "#440154"),
                                                    "Cell subtype" = c("dnT" = "#440154", "Treg" = "#46327e", "Proliferating" = "#1fa187", "TCM" = "#277f8e",   "TEM" = "#365c8d", "Naive" = "#4ac16d", "MAIT" = "#a0da39", "CTL" = "#fde725", "Trm" = "black"),
                                                    "Feature" = c("CDR3 length (AAs)" = "#21918c", "CDR3 hydrophobicity" = "#fde725", "Probability of generation" = "#440154"))),
        show_row_names = FALSE,
        column_title = "Clonotypes bins\n(% of repertoire across patients)",
        column_title_side = "bottom",
        heatmap_legend_param = list(title = "Wilcox padj val (-log10) for\nyes vs. no colitis irAE"),
        layer_fun = function(j, i, x, y, width, height, fill) {
          v = pindex(mat_downsampled, i, j)
          l = v > -log10(0.05) & v < -log10(0.01)
          if (any(l)) {
            grid.text("*", x[l], y[l], gp = gpar(fontsize = 10))
          }
          q = v > -log10(0.01)
          if (any(q)) {
            grid.text("**", x[q], y[q], gp = gpar(fontsize = 10))
          }
        }
      )

```

# Perform feature comparisons instead between cell types rather than between irAE groups

```{r}

tra_cell_type_features <- combined_data %>%
  filter(Chain == "TRA") %>%
  mutate(is_CD4_prolif = if_else(predicted.celltype.l2 == "CD4 Proliferating", "Yes", "No"),
         is_CD4_TEM = if_else(predicted.celltype.l2 == "CD4 TEM", "Yes", "No"),
         is_CD8_naive = if_else(predicted.celltype.l2 == "CD8 Naive", "Yes", "No"),
         is_CD8_TCM = if_else(predicted.celltype.l2 == "CD8 TCM", "Yes", "No"),
         is_mait = if_else(predicted.celltype.l2 == "MAIT", "Yes", "No")) %>%
  group_by(patient, cdr3, Chain, v_gene, j_gene) %>%
  slice(1) %>%
  ungroup() %>%
  select(barcode, Chain, cdr3, pgen, cdr3_length, cdr3_hydrophobicity, group, predicted.celltype.l2, is_CD4_TEM, is_mait)

tra_cell_type_features

ggplot(tra_cell_type_features, aes(x = is_CD4_TEM, y = log10(pgen))) +
  geom_boxplot() +
  labs(x = "CD4 TEM status", y = "pgen (-log10)")

ggplot(tra_cell_type_features, aes(x = is_CD4_TEM, y = cdr3_length)) +
  geom_boxplot() +
  labs(x = "CD4 TEM status", y = "CDR3 length (AAs)")

ggplot(tra_cell_type_features, aes(x = is_mait, y = log10(pgen))) +
  geom_boxplot() +
  labs(x = "MAIT status", y = "pgen (-log10)")

ggplot(tra_cell_type_features, aes(x = is_mait, y = cdr3_length)) +
  geom_boxplot() +
  labs(x = "MAIT status", y = "CDR3 length (AAs)")

# cdr3 length TRA padj
# CD4 TEM: 4.5850e-04
# CD8 naive: 1.2900e-02
# CD8 TCM: 5.4400e-03
# MAIT: 2.3965e-10

# cdr3 pgen TRA padj
# MAIT: 6.635000e-13

#print(wilcox_result <- wilcox.test(tra_cell_type_features$cdr3_length ~ tra_cell_type_features$is_CD4_prolif, exact = FALSE))

# cdr3 length TRB padj
# CD8 naive: 0.001790

#print(p.adjust(c(0.2391, 0.9648, 0.000358, 0.1884, 0.08194), method = "BH"))

# cdr3 hydrophobicity TRB padj
# CD8 naive: 0.0044400
# CD8 TCM: 0.0026015

```