---
title: "scRNAseq reveals distinct T cell populations in irAE of ICI"
output: html_notebook
---

 - data reproduction of https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9873824/

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load necessary libraries

```{r}

library(Seurat)
library(Matrix)
library(patchwork)
library(ggplot2)
library(dplyr)

```

# Set theme

```{r}

theme_set(
  theme_bw(20) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_rect(colour="black", fill = NA, size = 1),
          axis.text = element_text(colour="black"),
          axis.ticks = element_line(colour="black"),
          axis.text.x = element_text(angle=0),
          legend.title = element_text(size = 10),
          legend.text = element_text(size = 7)))

```

# Read in data

```{r}

blood_t_cells_filtered <- readRDS("./saved_data/blood_t_cells_filtered.rds")

```

# Normalize data

```{r}

blood_t_cells_filtered <- NormalizeData(blood_t_cells_filtered)

```

# Select variable features

```{r}

blood_t_cells_filtered <- FindVariableFeatures(blood_t_cells_filtered, nfeatures = 2000)

```

# Identify the 10 most highly variable genes

```{r}

top10 <- head(VariableFeatures(blood_t_cells_filtered), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(blood_t_cells_filtered) + 
  theme(legend.position="top")
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE) + 
  theme(legend.position="none")
plot1 + plot2

```

# Scale the data

```{r}

blood_t_cells_filtered <- ScaleData(blood_t_cells_filtered)

```

# Linear dimensionality reduction using principal component analysis (PCA)

```{r}

blood_t_cells_filtered <- RunPCA(blood_t_cells_filtered, features = VariableFeatures(object = blood_t_cells_filtered), npcs = 50)

```

# Visualize contribution to variance from each PC

```{r}

ElbowPlot(blood_t_cells_filtered, ndims = ncol(Embeddings(blood_t_cells_filtered, "pca")))

```

# Visualize many genes contributing to PCs for first two PCs

```{r}

VizDimLoadings(blood_t_cells_filtered, dims = 1, reduction = "pca")

VizDimLoadings(blood_t_cells_filtered, dims = 2, reduction = "pca")

```

# Visualize top genes contributing most to each PC for many PCs

```{r}

PCHeatmap(blood_t_cells_filtered, dims = 1:20, cells = 500, balanced = TRUE, ncol = 4)

```

# Visualize PCA

```{r}

DimPlot(blood_t_cells_filtered, group.by = "orig.ident", reduction = "pca", raster = FALSE)

```

# Cluster the cells

```{r}

blood_t_cells_filtered <- FindNeighbors(blood_t_cells_filtered, dims = 1:20)

blood_t_cells_filtered <- FindClusters(blood_t_cells_filtered, resolution = 0.5)

```

# Non-linear dimension reduction for visualization

```{r}

blood_t_cells_filtered <- RunUMAP(blood_t_cells_filtered, dims = 1:20)

```

DimPlot(pbmc, reduction = "umap", label = TRUE)

# Visualize UMAP with patient samples annotated

```{r}

plot1 <- UMAPPlot(blood_t_cells_filtered, raster = FALSE)

plot1

```

# Visualize UMAP with cell clusters numbered

```{r}

plot1 <- DimPlot(blood_t_cells_filtered, reduction = "umap", label = TRUE, raster = FALSE)

plot1

```

# Use established markers to determine identities of cell clusters

```{r}

# take markers from https://www.cell.com/cms/10.1016/j.xcrm.2022.100868/attachment/c21c9923-ade8-4a28-8518-aa1f7bc0d1e5/mmc1 table S2

```

# Create KNetL plot

```{r}

# use zoom of 500 (as they used)

```

# Incorporate CITE-seq data

```{r}



```